---
title: "Describe trends in your data"
author: "Daniela Palleschi"
institute: Humboldt-Universität zu Berlin
footer: "Data Visualisation"
lang: en
date: 2023-04-13
format:
  revealjs: 
    output-file: slides-data_communication.html
    theme: [dark]
    width: 1600
    height: 900
    progress: true
    # smaller: true
    scrollable: true
    slide-number: c/t
    code-link: true
    code-overflow: wrap
    code-tools: true
    # logo: logos/hu_logo.png
    # css: logo.css
    incremental: true
    # number-sections: true
    toc: true
    toc-depth: 1
    toc-title: 'Topics'
    navigation-mode: linear
    controls-layout: bottom-right
    fig-cap-location: top
    font-size: 0.6em
    slide-level: 4
    chalkboard: true
    title-slide-attributes: 
      data-background-image: logos/logos.tif
      data-background-size: 15%
      data-background-position: 50% 92%
  html:
    self-contained: true
    output-file: sheet-data_communication.html
    number-sections: true
    toc: true
    code-overflow: wrap
    code-tools: true
  pdf:
    output-file: pdf-data_communication.pdf
    toc: true
    number-sections: false
    colorlinks: true
    code-overflow: wrap
editor_options: 
  chunk_output_type: console
bibliography: references.json
csl: references/apa.csl
---

```{r}
knitr::opts_chunk$set(eval = T, # evaluate chunks
                      echo = T, # 'print code chunk?'
                      message = F, # 'print messages (e.g., warnings)?'
                      error = F, # stop when error encountered
                      warning = F) # don't print warnings
```

```{r, eval = T, cache = F}
# Create references.json file based on the citations in this script
# make sure you have 'bibliography: references.json' in the YAML
# rbbt::bbt_update_bib("_data_communication.qmd")
```

# Load in dataset

```{r}
# load tidyverse
library(tidyverse)

# load data
df_lifetime <- readr::read_csv(here::here("data/tidy_data_lifetime_pilot.csv"), 
                               # for special characters
                               locale = readr::locale(encoding = "latin1") 
                               ) |>
  mutate_if(is.character,as.factor) |> # all character variables as factor
  filter(type == "critical", # only critical trials
         px != "px3", # this participant had lots of 0's for some reason
         region=="verb")
```

# Data collection

## Participants

Data was collected from `r length(unique(df_lifetime$px))` participants.
Data was collected from native speakers of English (*n* = `r length(unique(df_lifetime$px))`, female = `r length(unique(df_lifetime$px[df_lifetime$gender=="female"]))`).

# Results

## Reading times

### First-fixation time

Useful functions for descriptive statistics:

- `mean(x)`: mean of x
- `sd(x)`: standard deviation of x
- `round(x,3)`: round x to 3 decimal points

```{r}
# save overall mean and sd as objects
mean_ff <- mean(df_lifetime$ff)
sd_ff <- sd(df_lifetime$ff)
```

The mean first-fixation time at the critical verb region was `r round(mean_ff,1)` milliseconds (ms), and the standard deviation was `r round(sd_ff,1)` ms.

### Main effect of lifetime (dead vs. living)

```{r}
#| echo: false
summary_ff_lifetime <- df_lifetime |> 
  filter(region=="verb") |> 
  group_by(lifetime) %>%
  summarise(N = n(),
            mean.ff = mean(ff, na.rm = T),
            sd = sd(ff, na.rm = T))

knitr::kable(summary_ff_lifetime, digits=1,
             caption = "Table with summmary statistics for first-fixation duration at the verb region")
```

The living condition elicited similar, but slightly shorter, first fixation times (mean = `r round(summary_ff_lifetime$mean.ff[summary_ff_lifetime$lifetime=="living"],1)`, sd = `r round(summary_ff_lifetime$sd[summary_ff_lifetime$lifetime=="living"],1)`) than the dead condition (mean = `r round(summary_ff_lifetime$mean.ff[summary_ff_lifetime$lifetime=="dead"],1)`, sd = `r round(summary_ff_lifetime$sd[summary_ff_lifetime$lifetime=="dead"],1)`) conditions.


### Main effect of tense (PP vs. SF)


```{r}
#| echo: false
#| 
summary_ff_tense <- df_lifetime |> 
  filter(region=="verb") |> 
  group_by(tense) %>%
  summarise(N = n(),
            mean.ff = mean(ff, na.rm = T),
            sd = sd(ff, na.rm = T))

knitr::kable(summary_ff_tense, digits=1,
             caption = "Table with summmary statistics for first-fixation duration at the verb region")
```

The Simple Future condition (mean = `r round(summary_ff_tense$mean.ff[summary_ff_tense$tense=="SF"],1)`, sd = `r round(summary_ff_tense$sd[summary_ff_tense$tense=="SF"],1)`) elicited similar, but slightly shorter, first fixation times than the Present Perfect condition (mean = `r round(summary_ff_tense$mean.ff[summary_ff_tense$tense=="PP"],1)`, sd = `r round(summary_ff_tense$sd[summary_ff_tense$tense=="PP"],1)`).

### Nested effects (lifetime within each level of tense)

```{r}
summary_ff <- df_lifetime |> 
  filter(region=="verb") |> 
  group_by(condition,lifetime,tense) %>%
  summarise(N = n(),
            mean.ff = mean(ff, na.rm = T),
            sd = sd(ff, na.rm = T))

knitr::kable(summary_ff, digits=1,
             caption = "Table with summmary statistics for first-fixation duration at the verb region")
```

In the present perfect condition, the dead (mean = `r round(summary_ff$mean.ff[summary_ff$condition=="deadPP"],1)`, sd = `r round(summary_ff$sd[summary_ff$condition=="deadPP"],1)`) elicited longer reading times than the living (mean = `r round(summary_ff$mean.ff[summary_ff$condition=="livingPP"],1)`, sd = `r round(summary_ff$sd[summary_ff$condition=="livingPP"],1)`) conditions.

Alternatively:

```{r}
deadPP_ff <- df_lifetime |> 
  filter(region=="verb",
         condition=="deadPP") |> 
  summarise(N = n(),
            mean.ff = mean(ff, na.rm = T),
            sd = sd(ff, na.rm = T))
```

`r deadPP_ff$mean.ff`

- number of observations, number of participants

- printing tables
- in-line code

# Describing plots

## Main effect of lifetime

```{r}
#| echo: false
plot_inter_lifetime <- df_lifetime |> 
  filter(region == "verb") |> 
  ggplot(aes(x = lifetime, y = ff, 
                     shape = lifetime,
                     group = lifetime,
                     color = lifetime)) +
  labs(title="Mean first-fixation time by lifetime",
       x = "Lifetime",
       y = "First fix (ms)",
       shape = "Lifetime", group = "Lifetime", color = "Lifetime", linetype = "Lifetime") +
    stat_summary(fun = "mean", geom = "line", position = position_dodge(0.2), aes(group=1), color = "black") +
  stat_summary(fun = "mean", geom = "point", size = 3, position = position_dodge(0.2)) +
  stat_summary(fun.data = "mean_cl_normal", geom = "errorbar", width = .2
               , position = position_dodge(0.2)) +
  theme_bw()
```

## Main effect of tense

```{r}
#| echo: false
plot_inter_tense <- df_lifetime |> 
  filter(region == "verb") |> 
  ggplot(aes(x = tense, y = ff, 
                     shape = tense,
                     group = tense,
                     color = tense)) +
  labs(title="Mean first-fixation time by tense",
       x = "Tense",
       y = "First fix (ms)",
       shape = "Tense", group = "Tense", color = "Tense", linetype = "Tense") +
    stat_summary(fun = "mean", geom = "line", position = position_dodge(0.2), aes(group=1), color = "black") +
  stat_summary(fun = "mean", geom = "point", size = 3, position = position_dodge(0.2)) +
  stat_summary(fun.data = "mean_cl_normal", geom = "errorbar", width = .2
               , position = position_dodge(0.2)) +
  theme_bw()
```

## Nested effects

```{r}
#| label: fig-inter-conditions

plot_inter_condition <- df_lifetime |> 
  filter(region == "verb") |> 
  ggplot(aes(x = lifetime, y = ff, 
                     shape = tense,
                     group = tense,
                     color = tense)) +
  labs(title="Mean first-fixation time by condition",
       x = "Lifetime",
       y = "First fix (ms)",
       shape = "Tense", group = "Tense", color = "Tense", linetype = "Tense") +
  stat_summary(fun = "mean", geom = "point", size = 3, position = position_dodge(0.2)) +
  stat_summary(fun = "mean", geom = "line", position = position_dodge(0.2), aes(linetype=tense)) +
  stat_summary(fun.data = "mean_cl_normal", geom = "errorbar", width = .2
               , position = position_dodge(0.2)) +
  theme_bw()
```

```{r}
ggpubr::ggarrange(plot_inter_lifetime,
                  plot_inter_tense,
                  plot_inter_condition,
                  nrow = 1, # keep plots in a single row
                  labels = c("A","B","C") # label plots A B and C
                  )
```


The figure shows the mean first fixation times for each condition.

@winter_statistics_2019

As stated in @winter_linear_2013, linear models are models.

# Session Info

```{r}
sessionInfo()
```


# References {.unlisted .unnumbered visibility="uncounted"}

::: {#refs custom-style="Bibliography"}
:::

