---
title: "Data description"
subtitle: "Reproducible Analyses (SoSe23)"
author: "Daniela Palleschi"
institute: Humboldt-Universität zu Berlin
lang: en
date: 2023-04-12
format:
  html:
    output-file: sheet-data_description.html
    toc: true
    toc-depth: 2
    toc-title: 'Overview'
    code-link: true
    code-overflow: wrap
    code-tools: true
    number-sections: true
    self-contained: true
  pdf:
    output-file: pdf-data_description.pdf
    toc: true
    number-sections: false
    colorlinks: true
    code-overflow: wrap
editor_options: 
  chunk_output_type: console
---

```{r, echo = F}
knitr::opts_chunk$set(eval = T, # evaluate = T for REPRODUCIBLE analyses
                      echo = F, # 'print code chunk?'
                      message = F, # print messages?
                      error = T, # render even if errors encountered?
                      warning = F) # print warnings?
```

# Lifetime pilot data

The following table gives a brief description of the variables (columns) from the dataset `data/data_lifetime_pilot.csv`.

```{r}
#| echo: false
#| eval: true

library(tidyverse)

# this was copied from my_eyes_only
vars <-
  c(
    RECORDING_SESSION_LABEL = "participant ID",
    TRIAL_INDEX = "trial number",
    EYE_USED = "which eye was tracked",
    IA_DWELL_TIME = "total reading time",
    IA_FIRST_FIXATION_DURATION = "first fixation duration",
    IA_FIRST_RUN_DWELL_TIME = "first-pass reading time",
    IA_FIXATION_COUNT = "number of fixations",
    IA_ID = "ROI number",
    IA_LABEL = "ROI text",
    IA_REGRESSION_IN = "whether the ROI had any regression(s) in",
    IA_REGRESSION_IN_COUNT = "number of regressions in",
    IA_REGRESSION_OUT = "whether the ROI had any regression(s) out",
    IA_REGRESSION_OUT_COUNT = "number of regressions out",
    IA_REGRESSION_PATH_DURATION = "regression path duration",
    KeyPress = "binary naturalness judgement key press value (4 or 5)",
    rt = "reaction time from beginning of critical sentence to button press",
    bio = "lifetime-context sentence text",
    critical = "critical sentence text",
    gender = "referent gender",
    item_id = "item number",
    list = "list number",
    match = "whether filler items matched or not",
    condition = "critical condition (or 'filler' for fillers)",
    name = "referent name",
    name_vital_status = "referent lifetime",
    tense = "tense (Present Perfect or Simple Future, or 'filler' for fillers)",
    type = "critical or filler",
    yes_press = "the key press value that corresponded to 'yes' in the experimental set-up. This was counterbalanced between participants (left-key: 4, right-key: 5)"
  )
```

```{r}
#| echo: false

# print vars (as data frame)

knitr::kable(data.frame(vars),
             col.names = "Description",
             caption = "Variable names and descriptions for dataset 'data\\_lifetime\\_pilot.csv'") %>% 
  kableExtra::kable_styling(full_width = T,
                            font_size = 12)
```

# Tidy lifetime pilot data

After our Data Wrangling session, your data should have the following variables (with different names, if you prefer) after our data wrangling session. This should be saved as `data_tidy_lifetime_pilot.csv` (or another name if you prefer).

```{r}
#| echo: false
#| eval: true
vars <-
  c(
    px = "participant ID",
    trial = "trial number",
    region = "ROI name",
    region_n = "ROI number",
    region_text = "ROI text",
    eye = "which eye was tracked",
    ff = "first fixation duration",
    fp = "first-pass reading time",
    rpd = "regression path duration",
    tt = "total reading time",
    fix_count = "number of fixations",
    reg_in = "whether the ROI had any regression(s) in",
    reg_in_count = "number of regressions in",
    reg_out = "whether the ROI had any regression(s) out",
    reg_out_count = "number of regressions in",
    rt = "reaction time from beginning of critical sentence to button press",
    bio = "lifetime-context sentence text",
    critical = "critical sentence text",
    gender = "referent gender",
    item_id = "item number",
    list = "list number",
    match = "whether filler items matched or not",
    condition = "critical condition (or 'filler' for fillers)",
    name = "referent name",
    lifetime = "referent lifetime",
    tense = "tense (Present Perfect or Simple Future, or 'filler' for fillers)",
    type = "critical or filler",
    yes_press = "the key press value that corresponded to 'yes' in the experimental set-up. This was counterbalanced between participants (left-key: 4, right-key: 5)",
    KeyPress = "binary naturalness judgement key press value (4 or 5)",
    accept = "whether KeyPress equaled the participant's 'yes' button, meaning a 'yes' naturalness judgement",
    accuracy = "whether 'accept' matched the 'congruence' level",
    px_accuracy = "participant's average accuracy across all items (critical and filler)"
  )
```

```{r}
#| echo: false

# print vars (as data frame)
knitr::kable(data.frame(vars),
             col.names = "Description",
             caption = "Variable names and descriptions for dataset `data/data\\_tidy\\_lifetime\\_pilot.csv`") %>% 
  kableExtra::kable_styling()
```

