---
title: "Model assumptions"
subtitle: "assumptions"
author: "Daniela Palleschi"
institute: Humboldt-Universität zu Berlin
footer: "Model assumptions"
lang: en
date: 2023-04-13
format:
  revealjs: 
    output-file: slides-assumptions.html
    theme: [dark]
    width: 1600
    height: 900
    progress: true
    # smaller: true
    scrollable: true
    slide-number: c/t
    code-link: true
    code-overflow: wrap
    code-tools: true
    code-annotations: below
    # logo: logos/hu_logo.png
    # css: logo.css
    incremental: true
    # number-sections: true
    toc: true
    toc-depth: 1
    toc-title: 'Topics'
    navigation-mode: linear
    controls-layout: bottom-right
    fig-cap-location: top
    font-size: 0.6em
    slide-level: 4
    title-slide-attributes: 
      data-background-image: logos/logos.tif
      data-background-size: 15%
      data-background-position: 50% 92%
    self-contained: true
  html:
    self-contained: true
    output-file: sheet-assumptions.html
    theme: [dark]
    number-sections: true
    toc: true
    code-overflow: wrap
    code-tools: true
  pdf:
    output-file: pdf-assumptions.pdf
    toc: true
    number-sections: false
    colorlinks: true
    code-overflow: wrap
editor_options: 
  chunk_output_type: console
bibliography: references/references.json
csl: references/apa.csl
---

```{r, echo = T}
knitr::opts_chunk$set(eval = T, # change this to 'eval = T' to reproduce the analyses; make sure to comment out
                      echo = T, # 'print code chunk?'
                      message = F, # 'print messages (e.g., warnings)?'
                      error = F,
                      warning = F)
```

```{r}
#| echo: false
## play sound if error encountered
### from: https://sejohnston.com/2015/02/24/make-r-beep-when-r-markdown-finishes-or-when-it-fails/
options(error = function(){    # Beep on error
  beepr::beep(sound = "wilhelm")
  Sys.sleep(2) # 
  }
 )
## and when knitting is complete
.Last <- function() {          # Beep on exiting session
  beepr::beep(sound = "ping")
  Sys.sleep(6) # allow to play for 6 seconds
  }
```

```{r, eval = T, cache = F}
#| echo: false
# Create references.json file based on the citations in this script
# make sure you have 'bibliography: references.json' in the YAML
# rbbt::bbt_update_bib("_assumptions.qmd")
```

# Learning objectives {.unnumbered}

Today we will learn...

- how to check model assumptions
- how to transform our data to meet model assumptions

# Resources {.unnumbered}

Sections 6.3, 6.4 and 7.9  in @winter_statistics_2019

# Set-up {.unnumbered}

1. Start with a clean R Environment (Session > Restart R).

2. Suppress scientific notation (make very small numbers easier to read):

```{r}
# suppress scientific notation
options(scipen=999)
```


## Packages {.unnumbered}

```{r}
pacman::p_load(tidyverse,
               broom,
               patchwork,
               sjPlot,
               knitr,
               kableExtra,
               Rmisc)
```

## Data {.unnumbered}

  - force character variables to factors
  - filter for the verb region from critical items only, remove participant 3, and remove values of first-fixtation that are

```{r}
# load in dataset
df_crit_verb <-
  readr::read_csv(
    here::here("data/tidy_data_lifetime_pilot.csv"),
    # for special characters
    locale = readr::locale(encoding = "latin1")
  ) |>
  mutate_if(is.character, as.factor) |> # all character variables as factor
  mutate(lifetime = fct_relevel(lifetime, "living", "dead"),
         tense = fct_relevel(tense, "PP", "SF")) |>
  filter(type == "critical", # only critical trials
         px != "px3", # px3 had a lot of missing values
         fp > 0, # only values of fp above 0
         region == "verb") %>% # critical region only
  droplevels() # remove any factor levels with no observations
```

# Review: (multiple) linear regression

- when we run a linear model, we are fitting a line (predicted values) to our data (observed values)
  + the *intercept* is the ***predicted*** value of *y* (our outcome) when *x* (our predictor) is 0
  + the *slope* is the ***predicted*** change in *y* with an increase of 1-unit of *x*
  + the *residuals* measure the ***difference*** between our ***predicted values*** of *y* from the ***observed values*** of *y*

$$
\begin{align}
observed\;values &= fitted\;values + residuals\\
residuals &= observed\;values - fitted\;values
\end{align}
$$
  
## Our model

### Check contrasts

```{r}
#| output-location: column-fragment
contrasts(df_crit_verb$lifetime)
```

```{r}
#| output-location: column-fragment
contrasts(df_crit_verb$tense)
```

### Re-order predictor levels

```{r}
# set contrasts
df_crit_verb <- df_crit_verb %>% 
  mutate(lifetime = fct_relevel(lifetime, "living", "dead"),
         tense = fct_relevel(tense, "PP", "SF"))
```

### Set sum coding

Set contrasts

```{r}
#| output-location: column
contrasts(df_crit_verb$lifetime) <- c(-0.5, +0.5)
contrasts(df_crit_verb$tense) <- c(-0.5, +0.5)
```

Check contrasts

```{r}
#| output-location: column-fragment
contrasts(df_crit_verb$lifetime)
```


```{r}
#| output-location: column-fragment
contrasts(df_crit_verb$tense)
```

### Run model

```{r}
# fit linear model
fit_fp <- df_crit_verb %>%
  lm(fp ~ lifetime*tense, data = .)
```

### Print summary

```{r}
summary(fit_fp)
```

# Assumptions

- refer to ***residuals***
  + i.e., the difference between the observed and the fitted (predicted) values

1.   *normality assumption*
      -   residuals of the model are (approximately) normally distributed
2.   *constant variance assumption* (homoscedasticity)
      -   spread of residuals should be (approximately) equal along the regression line
3.   absence of *collinearity*
      -   predictors should not be correlated with each other
4.   *independence*
      -   data points should be independent from each other

## Checking assumptions

- we typically assess **normality** and **homoscedasticity** visually with histograms, Q-Q plots, and residual plots
- we can assess collinearity with VIFs
- the assumption of independence is dealt with conceptually

::: {.content-visible when-format="revealjs"}
```{r echo = F, fig.align = "center"}
#| fig-cap: "Image source: @winter_statistics_2019 (all rights reserved)"
#| fig-cap-location: bottom

# invert colours for dark mode in slides
library(magick)
y <- magick::image_read(here::here("media/Winter_2019_assumptions_plots.png"))

magick::image_negate(y)
```
:::

::: {.content-hidden when-format="revealjs"}
```{r echo = F, fig.align = "center"}
#| fig-cap: "Image source: @winter_statistics_2019 (all rights reserved)"
#| fig-cap-location: bottom
magick::image_read(here::here("media/Winter_2019_assumptions_plots.png"))
```
:::

# Constant variance: Residual plot

- doesn't look like a blob with categorical data, but rather stripes
  + what's important is that the distribution (vertically) is roughly similar across stripes
  + why are predicted values for categorical data stripey?
  
```{r}
#| echo: false
#| results: hide
df_crit_verb$predicted <- predict(fit_fp)   # Save the predicted values
df_crit_verb$residuals <- residuals(fit_fp) # Save the residual values

# Quick look at the actual, predicted, and residual values
df_crit_verb %>% select(lifetime, tense, predicted, residuals) %>% head()
```

```{r}
#| echo: false
fit_fp %>% 
ggplot(aes(x = .fitted, y = .resid)) +
  labs(title = "Residual plot (fp ~ lifetime*tense)") +
  geom_point() +
  geom_hline(yintercept = 0) +
  theme_bw()  # Add theme for cleaner look
```

# Normality assumption

- can be inspected e.g., with a histogram or Q-Q plot

```{r}
df_crit_verb |> 
  filter(fp > 0) |> 
  mutate(half = if_else(trial >= 104, "1st","2nd")) |> 
  ggpubr::ggqqplot(x = "fp")
```

## Normality assumption {-}

- how about by participant and experimental half?

```{r}
df_crit_verb |> 
  filter(fp > 0) |> 
  mutate(half = if_else(trial >= 104, "1st","2nd")) |> 
  ggpubr::ggqqplot(x = "fp",
                    color = "half",
                    facet.by = "px")
```

## Normality assumption {-}

- normal distribution is symmetrical, are our residuals normally distributed?   

```{r}
#| echo: false
# plot(density(resid(fit_fp)))
fig_dens <- ggplot(data = df_crit_verb, aes(x = fit_fp$residuals)) +
    geom_density() +
    labs(title = 'Density of Residuals', x = 'Residuals', y = 'Frequency')
fig_hist <- ggplot(data = df_crit_verb, aes(x = fit_fp$residuals)) +
    geom_histogram() +
    labs(title = 'Histogram of Residuals', x = 'Residuals', y = 'Frequency')
```

```{r}
#| echo: false
#| label: fig-normality
#| fig-cap: Visualising normality of residuals
fig_dens + fig_hist
```




## Normality assumption {-}

- reading time data tends to be *positively skewed*
  + so the residuals also tend to be positively skewed
- data with a skewed distribution do not meet the normality assumption
- a fix: nonlinear transformations
  + the most common: the log transformation
- log-transforming your data makes larger numbers smaller (and small numbers smaller too)
  + the difference between smaller numbers and larger numbers shrinks
  + can make skewed data normally distributed

### Log transformation

-   for more see Section 5.4 in @winter_statistics_2019

-   the R funtion `log()` computes the 'natural logarithm' (and is the inverse of the exponential `exp()`)
  -   `log()` makes large numbers smaller
  -   `exp()` makes small numbers larger

```{r}
#| output-location: column-fragment
log(2)
```
```{r}
#| output-location: column-fragment
log(1:10)
```
```{r}
#| output-location: column-fragment
log(c(10,20,30,40,100))
```

```{r}
#| output-location: column-fragment
exp(1:10)
```
```{r}
#| output-location: column-fragment
exp(log(1:10))
```

## Fit model (log-transformed)

-   continuous variables truncated at 0 typically have a *positive skew*
    -   a lot of small values (e.g., `tt` \< 500ms), with some larger values (\> `tt` 1000)
    -   this usually means our residuals are also positively skewed, i.e., not normally distributed
-   so we typically log-transform raw reading/reaction times for our linear models

```{r}
#| output-location: column-fragment
# fit simple linear model with log
fit_fp_log <- df_crit_verb %>%
  filter(fp > 0) %>% # important! you can't log transform 0
  lm(log(fp) ~ lifetime*tense, data = .)
summary(fit_fp_log)
```

### Check assumptions

:::: columns
::: {.column width="50%"}
Raw first-pass reading times
```{r}
plot(density(resid(fit_fp)))
```
:::

::: {.column width="50%"}
Log first-pass reading times
```{r}
plot(density(resid(fit_fp_log)))
```
:::
::::

### Check assumptions {-}

:::: columns
::: {.column width="50%"}
Raw first-pass reading times
```{r}
qqnorm(residuals(fit_fp))
qqline(residuals(fit_fp), col="red")
```
:::

::: {.column width="50%"}
Log first-pass reading times
```{r}
qqnorm(residuals(fit_fp_log))
qqline(residuals(fit_fp_log), col="red")
```
:::
::::

### Check assumptions: constant variance {-}

:::: columns
::: {.column width="50%"}
Raw first-pass reading times
```{r}
fit_fp %>% 
ggplot(aes(x = .fitted, y = .resid)) +
  labs(title = "Residual plot (fp ~ lifetime*tense)") +
  geom_point() +
  geom_hline(yintercept = 0) +
  theme_bw()  # Add theme for cleaner look
```
:::

::: {.column width="50%"}
Log first-pass reading times
```{r}
fit_fp_log %>% 
ggplot(aes(x = .fitted, y = .resid)) +
  labs(title = "Residual plot (log(fp) ~ lifetime*tense)") +
  geom_point() +
  geom_hline(yintercept = 0) +
  theme_bw()  # Add theme for cleaner look
```
:::
::::

### Approaching normality: log

- so it seems like the log-transformed values have made the data more normal

## Interpreting log

```{r}
augment(fit_fp_log, data = df_crit_verb[df_crit_verb$fp > 0,]) %>% 
  distinct(lifetime,tense,.fitted) %>% 
  arrange(lifetime) 
```

```{r}
df_crit_verb %>% 
  mutate(predicted_raw = predict(fit_fp),
         predicted_log = predict(fit_fp_log)) %>% 
  distinct(lifetime,tense,predicted_raw, predicted_log) %>% 
  arrange(lifetime) 
```

---

```{r}
exp(coef(fit_fp_log)['(Intercept)'])
coef(fit_fp)['(Intercept)']
```

- the exponential (back-transformation) of the log slopes correspond to the change in *percentage*
```{r}
(exp(coef(fit_fp_log)['lifetime1'])-1)*100
```

```{r}
(exp(coef(fit_fp_log)['tense1'])-1)*100
```

```{r}
# living-SF
coef(fit_fp)['(Intercept)'] + (coef(fit_fp)['lifetime1'] * -0.5)

exp(coef(fit_fp_log)['(Intercept)']) * (exp(coef(fit_fp_log)['lifetime1']* -0.5))
exp(coef(fit_fp_log)['(Intercept)']) * (exp(coef(fit_fp_log)['lifetime1']* +0.5))

```
```{r}
# living-SF
exp(coef(fit_fp_log)['(Intercept)']) * exp(coef(fit_fp_log)['lifetime1'] * -0.5 + coef(fit_fp_log)['tense1'] * -0.5 )
```
```{r}
# dead-PP
exp(coef(fit_fp_log)['(Intercept)']) * exp(coef(fit_fp_log)['lifetime1'] * +0.5 + coef(fit_fp_log)['tense1'] * -0.5 )
```
```{r}
# dead-SF
coef(fit_fp_log)['(Intercept)'] + coef(fit_fp_log)['lifetime1'] * 0.5 + coef(fit_fp_log)['tense1'] * 0.5 
```


# Collinearity

- your predictors should not be correlated
  - this can be checked using the 'variance inflation factors' (VIFs), which measure the degree to which a predictor can be accounted for by other predictors
- there are different guidelines for appropriate cut-offs for VIFs, with some considering a VIF >10 indicates colliniearity
  
```{r}
pacman::p_load(car)
```

```{r}
vif(fit_fp)
```

- importantly, sample size interactions with collinearity (the more data you have, the more precisely you can measure even in the face of collinearity)
  + so, an alternative to dropping a predictor is to collect more data (if possible)

# Independence assumption

- important in *repeated measures* design, where we collect multiple data points across conditions per participant
- the importance of the indepdence assumption far outweighs that of the other assumptions
- having data points that are linked, but not telling your model this information, can massively inflate your Type I error (probability of finding a false positive)

## What is (in)dependence of data?

  + e.g., each roll of a die is independent
    + that is, if you roll a 6, this has no bearing on what you will roll next
    + if you roll a die 10 times and I roll a die 10 times, the values we get will not be linked to who rolled the die
  + reading times are dependent
    + if we collect eye-tracking reading times from two people, the data points from each person will be dependent
      + this is because each person will have their own reading pace, so their data points will tend to cluster
    + also true for experimental *item*
      + some items will likely elicit a stronger effect than others

## Check independence

- independence is a conceptual consideration, it can't be checked visually or numerically
- it is a question of experimental *design*
- a way to think about it: can my data be linked/clustered/grouped by some means?
  + e.g., yes, by *participant*, by *item*, even by *verb*
- there of course are many ways our data might cluster, contributing random error that would not be expected **to be repeated**
  + we have specific predictions for the effects of `lifetime` or `tense`, and would expect them to replicate if we were to re-run the study
  + but we cannot *predict* how certain participants will pattern, nor how certain items will pattern
  + these are ***random variables***, essentially error that we can try to explain by means of telling the model "check for effects by participant and item"

## Accounting for independence

- there is a very powerful tool that allows us to include dependence in our models: linear mixed models (LMMs)
- LMMs are *mixed* because they include *fixed effects* (`lifetime`, `tense`), and *random effects* (in eye-tracking experiments, typically `participant`, `item`)
  - also called linear mixed *effects* models (LMEMs), or multilevel models (or hierarchical models in some cases)

# `performance` package

- the `performance` package has a nice function `check_model()` that produces 6 plots to check model fit

```{r}
#| label: fig-performance-log
#| fig-cap: Performance of model with log reading times
performance::check_model(fit_fp_log)
```

---

```{r}
#| label: fig-performance-raw
#| fig-cap: Performance of model with log reading times
performance::check_model(fit_fp)
```


# Communicating your results

-   model summaries can be provided via tables and/or figures
    -   you should always report the t-values and p-values of an effect
-   this is where the `tidy()` function from `broom` is handy in combination with `knitr::kable(digits = x)` and `kableExtra::kable_styling()`

```{r}
tidy(fit_fp_log) %>% 
  kable(digits = 10,
        col.names = c("Coefficient",
                      "Estimate (log)",
                      "SE",
                      "t-value",
                      "p-value")) %>% 
  kable_styling()
```


## P-value formatting

- you can create a function the replaces p-values

```{r}
# source: https://stackoverflow.com/questions/37470202/in-line-code-for-reporting-p-values-001-in-r-markdown
# OR USE
pacman::p_load(broman) 
format_pval <- function(x){
  if (x < .001) return(paste('<', '.001'))
  if (x < .01) return(paste('<', '.01'))
  if (x < .05) return(paste('<', '.05'))
  paste('=', myround(x, 3))  # if above .05, print p-value to 3 decimalp points
}
```

```{r}
make_stars <- function(pval) {
  stars = ""
  if(pval <= 0.001)
    stars = "***"
  if(pval > 0.001 & pval <= 0.01)
    stars = "**"
  if(pval > 0.01 & pval <= 0.05)
    stars = "*"
  if(pval > 0.05 & pval <= 0.1)
     stars = "."
  stars
}
```

## Formatted table

```{r}
tidy(fit_fp_log) %>% 
  mutate(format = sapply(p.value, function(x) format_pval(x))) %>% 
  mutate(signif = sapply(p.value, function(x) make_stars(x))) %>%
  select(-p.value) %>% 
  kable(digits = 10,
        col.names = c("Coefficient",
                      "Estimate (log)",
                      "SE",
                      "t-value",
                      "p-value",
                      "sign")) %>% 
  kable_styling()
```

## Plots

```{r}
df_crit_verb |> 
  filter(fp > 0) |> 
  mutate(log_ff = log(fp)) |> 
  mutate(half = if_else(trial >= 104, "1st","2nd")) |> 
  ggpubr::ggqqplot(x = "log_ff")
```

### sjPlot

```{r}
#| echo: false
fig_sjplot_coef <- plot_model(fit_fp_log) +
  geom_hline(yintercept=0) +
  theme_bw()

fig_sjplot_int <- plot_model(fit_fp_log, type = "int") +
  geom_hline(yintercept=0) +
  geom_line(group=1, position = position_dodge(0.1)) +
  theme_bw()
```

```{r}
#| echo: false
#| label: fig-sjPlot
#| fig-label: Model summaries
fig_sjplot_coef + fig_sjplot_int + plot_annotation(tag_levels = "A")
```


### Distributions

```{r}
#| echo: false
fig_scatter <-
  df_crit_verb %>% 
  filter(fp > 0) %>% 
  ggplot(aes(x = tense, y = fp, colour = lifetime, shape = lifetime, fill = lifetime)) + 
  labs(title = "Distribution of first-pass reading times") +
  geom_point(position = position_jitterdodge(jitter.width = 0.3), size = 2, alpha = .5) +
  introdataviz::geom_split_violin(alpha = .4) +
  theme_bw()+
  theme(text = element_text(size=8))

fig_scatter_log <-
  df_crit_verb %>% 
  filter(fp > 0) %>% 
  ggplot(aes(x = tense, y = log(fp), colour = lifetime, shape = lifetime, fill = lifetime)) + 
  labs(title = "Distribution of first-pass reading times") +
  geom_point(position = position_jitterdodge(jitter.width = 0.3), size = 2, alpha = .5) +
  introdataviz::geom_split_violin(alpha = .4) +
  theme_bw()+
  theme(text = element_text(size=8))
```

```{r}
#| echo: false
#| label: fig-scatter
#| fig-label: Distribution of raw (A) and log (B) first-pass reading times
fig_scatter + fig_scatter_log + plot_annotation(tag_levels = "A")
```

### Errorbar

```{r}
fig_error <-
  df_crit_verb %>% 
  filter(fp > 0) %>% 
  summarySEwithin(measurevar="fp", withinvars=c("lifetime", "tense"), idvar="px") %>% 
  mutate(upper = fp+ci,
         lower = fp-ci) %>% 
  ggplot(aes(x = lifetime, y = fp, colour = tense, shape = tense)) + 
  labs(title = "Mean first-pass reading times (with 95% CIs)") +
  geom_point(position = position_dodge(0.2), size = 2) +
  geom_line(position = position_dodge(0.2), aes(group=tense)) +
  geom_errorbar(aes(ymin=lower,ymax=upper), position = position_dodge(0.2), width = .2) +
  theme_bw() +
  theme(text = element_text(size=8))

fig_error_log <-
  df_crit_verb %>% 
  filter(fp > 0) %>% 
  mutate(fp_log = log(fp)) %>% 
  summarySEwithin(measurevar="fp_log", withinvars=c("lifetime", "tense"), idvar="px") %>% 
  mutate(upper = fp_log+ci,
         lower = fp_log-ci) %>% 
  ggplot(aes(x = lifetime, y = fp_log, colour = tense, shape = tense)) + 
  labs(title = "Mean first-pass reading times (with 95% CIs)") +
  geom_point(position = position_dodge(0.2), size = 2) +
  geom_line(position = position_dodge(0.2), aes(group=tense)) +
  geom_errorbar(aes(ymin=lower,ymax=upper), position = position_dodge(0.2), width = .2) +
  theme_bw() +
  theme(text = element_text(size=8))
```

```{r}
#| label: fig-error
#| echo: false
#| fig-label: Mean raw (A) and log (B) first-pass reading times (95% CIs)
fig_error + fig_error_log + plot_annotation(tag_levels = "A")
```

# Summary

-   we looked at some assumptions of linear models and how to check them

-   we saw how to log-transform our data so that our residuals are normally distributed

-   we learned the importance of the independence assumption, which is violated by repeated-measures designs

-   next, we'll learn how to account for lack of independence between the data with mixed models

# References {.unlisted .unnumbered visibility="uncounted"}

::: {#refs custom-style="Bibliography"}
:::
