---
title: "Reproducible Workflow"
subtitle: "Loading and eye-balling a dataset"
author: "Daniela Palleschi"
institute: Humboldt-Universität zu Berlin
lang: en
date: 2023-04-12
format:
  revealjs: 
  # all docs:
    output-file: slide-workflow.html
  # revealjs and html:
    # toc
    # code
    toc: true
    toc-depth: 1
    toc-title: 'Overview'
    code-link: true
    code-overflow: wrap
    code-tools: true
    css: styles.css
  # revealjs specific:
    theme: [dark]
    width: 1600
    height: 900
    progress: true
    font-size: 0.6em
    slide-level: 4
    slide-number: c/t
    navigation-mode: linear
    # smaller: true
    incremental: true
    scrollable: true
    controls-layout: bottom-right
    fig-cap-location: top
    title-slide-attributes: 
      data-background-image: logos/logos.tif
      data-background-size: 15%
      data-background-position: 50% 92%
    # to add drawings:
    # chalkboard:
    #   src: drawings/day1-drawings.json
    embed-resources: true
    # logo: logos/hu_logo.png
    # css: logo.css
    # number-sections: true
  html:
    output-file: sheet-workflow.html
    embed-resources: true
    toc: true
    toc-depth: 2
    toc-title: 'Overview'
    code-link: true
    code-overflow: wrap
    code-tools: true
    number-sections: false
  pdf:
    output-file: pdf-workflow.pdf
    toc: true
    number-sections: false
    colorlinks: true
    code-overflow: wrap
editor_options: 
  chunk_output_type: console
bibliography: references/references.json
csl: references/apa.csl
---


# Set-up {.unlisted .unnumbered visibility="uncounted"}

```{r}
knitr::opts_chunk$set(eval = T, # evaluate = T for REPRODUCIBLE analyses
                      echo = T, # 'print code chunk?'
                      message = F, # print messages?
                      error = T, # render even if errors encountered?
                      warning = F) # print warnings?
```

```{r}
library(beepr) # beeps when code runs or fails
library(rbbt) # zotero plugin
library(magick) # to invert images
```

```{r}
#| code-fold: true
#| code-summary: beep code
## play sound if error encountered
### from: https://sejohnston.com/2015/02/24/make-r-beep-when-r-markdown-finishes-or-when-it-fails/
options(error = function(){    # Beep on error
  beepr::beep(sound = "wilhelm")
  Sys.sleep(2) # 
  }
 )
## and when knitting is complete
.Last <- function() {          # Beep on exiting session
  beepr::beep(sound = "ping")
  Sys.sleep(6) # allow to play for 6 seconds
  }
```

```{r, eval = T, cache = F}
#| code-fold: true
#| code-summary: rbbt code
#| 
# Create references.json file based on the citations in this script:
# 1. make sure you have 'bibliography: references/references.json' in the YAML
# 2. create a new folder called 'references'
# 3. run:
rbbt::bbt_update_bib("_workflow.qmd")
```

# Workflow bare minimum

- self-contained project
  + everything available in one folder
  + e.g., RProjects
- README file
  + a markdown (.md) file
  + describing the folder/analysis structure
  
::: footnote
@winter_statistics_2019
:::

# RProjects

- a folder containing
  + an `.RProj` file (which opens RStudio)
  + all folders/files required for a project
- `File > New Project > New Directory > New Project > New Project > Create Project`

## README

- to create an `.md` file: `File > New File > Markdown File`
- create informative heading
  + describe project purpose
  + describe folders/scripts as they currently are
- save/Preview as `README.md` in the project folder

## `here`

- `here` package
  + will always access the project folder
  + try running `here()` from within a project; what's the output?
  
## Project folder structure

::: {.content-visible when-format="revealjs"}
```{r eval = T, fig.env = "figure",out.width="100%", fig.align = "center", set.cap.width=T, fig.cap="Image source: @winter_statistics_2019"}
#| echo: false

# invert colours for dark mode in slides
magick::image_negate(magick::image_read(here::here("media/Winter_2019_project-folder.png")))
```
:::

::: {.content-visible when-format="html"}
```{r eval = T, fig.env = "figure",out.width="100%", fig.align = "center", set.cap.width=T, fig.cap="Image source: @winter_statistics_2019"}
#| echo: false

library(magick)
magick::image_read(here::here("media/Winter_2019_project-folder.png"))
```
:::

::: {.content-visible when-format="pdf"}
```{r eval = T, fig.env = "figure",out.width="100%", fig.align = "center", set.cap.width=T, fig.cap="Image source: @winter_statistics_2019"}
#| echo: false
library(magick)
magick::image_read(here::here("media/Winter_2019_project-folder.png"))
```
:::


# Reproducible code

- is located within a project
  + which also contains all relevant data/files
- runs linearly (from top to bottom)
  + loads all required packages at the top
  + uses file paths relative to its project
  + is created/edited after running `Session > Restart R`
- at the very least, ends with a section (e.g., `# Session Info`) containing `sessionInfo()`
  + but other options: `renv` package, `targets` package, `docker` for environment containers

# Checklist

::: {.column width="50%"}
**RProject**

  + `.RProj`
  + `README.md`
  + `data/`
  + `notes/` (a folder for your qmd files)
:::

::: {.column width="50%"}
**Scripts** (`.qmd`/`.Rmd`)

  + load libraries at beginning
  + `sessionInfo()` at the end
  + chunks run linearly (top-to-bottom)
:::

# Session Info

```{r}
#| code-fold: true
#| code-summary: "Show code"
sessionInfo()
```

# References {.unlisted .unnumbered visibility="uncounted"}

::: {#refs custom-style="Bibliography"}
:::