---
title: "Model assumptions"
subtitle: "Refresher"
author: "Daniela Palleschi"
institute: Humboldt-Universität zu Berlin
footer: "Model assumptions"
lang: en
date: 2023-04-13
format:
  revealjs: 
    output-file: slides-refresher.html
    theme: [dark]
    width: 1600
    height: 900
    progress: true
    # smaller: true
    scrollable: true
    slide-number: c/t
    code-link: true
    code-overflow: wrap
    code-tools: true
    code-annotations: below
    # logo: logos/hu_logo.png
    # css: logo.css
    incremental: true
    # number-sections: true
    toc: true
    toc-depth: 1
    toc-title: 'Topics'
    navigation-mode: linear
    controls-layout: bottom-right
    fig-cap-location: top
    font-size: 0.6em
    slide-level: 4
    chalkboard: true
    title-slide-attributes: 
      data-background-image: logos/logos.tif
      data-background-size: 15%
      data-background-position: 50% 92%
  html:
    self-contained: true
    output-file: sheet-refresher.html
    theme: [dark]
    number-sections: true
    toc: true
    code-overflow: wrap
    code-tools: true
  pdf:
    output-file: pdf-refresher.pdf
    toc: true
    number-sections: false
    colorlinks: true
    code-overflow: wrap
editor_options: 
  chunk_output_type: console
bibliography: references/references.json
csl: references/apa.csl
---

```{r, echo = T}
knitr::opts_chunk$set(eval = T, # change this to 'eval = T' to reproduce the analyses; make sure to comment out
                      echo = T, # 'print code chunk?'
                      message = F, # 'print messages (e.g., warnings)?'
                      error = F,
                      warning = F)
```

```{r}
#| echo: false
## play sound if error encountered
### from: https://sejohnston.com/2015/02/24/make-r-beep-when-r-markdown-finishes-or-when-it-fails/
options(error = function(){    # Beep on error
  beepr::beep(sound = "wilhelm")
  Sys.sleep(2) # 
  }
 )
## and when knitting is complete
.Last <- function() {          # Beep on exiting session
  beepr::beep(sound = "ping")
  Sys.sleep(6) # allow to play for 6 seconds
  }
```

```{r, eval = T, cache = F}
#| echo: false
# Create references.json file based on the citations in this script
# make sure you have 'bibliography: references.json' in the YAML
rbbt::bbt_update_bib("_assumptions.qmd")
```

# Learning objectives

# Set-up

1. Start with a clean R Environment (Session > Restart R).

2. Suppress scientific notation (make very small numbers easier to read):

```{r}
# suppress scientific notation
options(scipen=999)
```

## Packages

```{r}
pacman::p_load(tidyverse,
               broom,
               patchwork,
               knitr,
               kableExtra)
```

## Set seed

- the `set.seed()` function is useful when generating random numbers in R
- it allws us to set a seed and generator for random number generation
- otherwise, every time I ran this code (and every time you ran code), I would get different random numbers
  - it therefore allows us to ensure full reproducibility

```{r}
set.seed(1954)
```

## Data

  - force character variables to factors
  - filter for the verb region from critical items only, remove participant 3, and remove values of first-fixtation that are

```{r}
# load in dataset
df_crit_verb <-
  readr::read_csv(
    here::here("data/tidy_data_lifetime_pilot.csv"),
    # for special characters
    locale = readr::locale(encoding = "latin1")
  ) |>
  mutate_if(is.character, as.factor) |> # all character variables as factor
  mutate(lifetime = fct_relevel(lifetime, "living", "dead"),
         tense = fct_relevel(tense, "PP", "SF")) |>
  filter(type == "critical", # only critical trials
         px != "px3", # px3 had a lot of missing values
         fp > 0, # only values of fp above 0
         region == "verb") %>% # critical region only
  droplevels() # remove any factor levels with no observations
```

# Review: linear regression

- when we run a linear model, we are fitting a line (predicted values) to our data (observed values)
  + the *intercept* is the ***predicted*** value of *y* (our outcome) when *x* (our predictor) is 0
  + the *slope* is the ***predicted*** change in *y* with an increase of 1-unit of *x*
  + the *residuals* measure the ***difference*** between our ***predicted values*** of *y* from the ***observed values*** of *y*
  
## Our model

1. Set contrasts

```{r}
#| output-location: column-fragment
contrasts(df_crit_verb$lifetime)
```

```{r}
# set contrasts
df_crit_verb <- df_crit_verb %>% 
  mutate(lifetime = fct_relevel(lifetime, "living", "dead"))
```

```{r}
contrasts(df_crit_verb$lifetime) <- c(-0.5, +0.5)
```

```{r}
#| output-location: column-fragment
contrasts(df_crit_verb$lifetime)
```


---

2. Run model

```{r}
# fit simple linear model
fit_fp_lifetime <- df_crit_verb %>%
  filter(fp > 0) %>%
  lm(fp ~ lifetime, data = .)
```

```{r}
summary(fit_fp_lifetime)
```

# Assumptions

- refer to ***residuals***
  + i.e., the difference between the observed and the fitted (predicted) values

-   *normality assumption*
    -   residuals of the model are (approximately) normally distributed
-   *constant variance assumption* (homoscedasticity)
    -   spread of residuals should be (approximately) equal along the regression line
-   absence of collinearity
-   independence

## Normality assumption

- can be inspected e.g., with a histogram or Q-Q plot

```{r}
df_crit_verb |> 
  filter(fp > 0) |> 
  mutate(half = if_else(trial >= 104, "1st","2nd")) |> 
  ggpubr::ggqqplot(x = "fp")
```

## Normality assumption {-}

- how about by participant and experimental half?

```{r}
df_crit_verb |> 
  filter(fp > 0) |> 
  mutate(half = if_else(trial >= 104, "1st","2nd")) |> 
  ggpubr::ggqqplot(x = "fp",
                    color = "half",
                    facet.by = "px")
```

## Normality assumption {-}

- density plot (of residuatls)

```{r}
plot(density(resid(fit_fp_lifetime)))
```

## Normality assumption {-}

- residual plot: not very helpful for binomial data!

```{r}
plot(fitted(fit_fp_lifetime),residuals(fit_fp_lifetime))
```

```{r}
par(mfrow = c(2, 2))  # Split the plotting panel into a 2 x 2 grid
plot(fit_fp_lifetime)
```

## Obtain predicted values

```{r}
df_crit_verb$predicted <- predict(fit_fp_lifetime)   # Save the predicted values
df_crit_verb$residuals <- residuals(fit_fp_lifetime) # Save the residual values

# Quick look at the actual, predicted, and residual values
df_crit_verb %>% select(lifetime, predicted, residuals) %>% head()
```

```{r}
pj = position_jitter(0.2)

df_crit_verb %>% 
ggplot(aes(x = lifetime, y = fp)) +
  geom_smooth(method = "lm", se = FALSE, color = "lightgrey", aes(group=1)) +  # Plot regression slope
  # geom_segment(aes(xend = lifetime, yend = predicted), alpha = .2) +  # alpha to fade lines
  geom_point() +
  geom_point(aes(y = predicted), shape = 17, colour = "red", size = 4) +
  theme_bw()  # Add theme for cleaner look
```




## Normality assumption {-}

- reading time data tends to be *positively skewed*
  + so the residuals also tend to be positively skewed
- data with a skewed distribution is does not meet the normality assumption
- a fix: nonlinear transformations
  + the most common: the log transformation
- log-transforming your data makes larger numbers smaller (and small numbers smaller too)
  + the difference between smaller numbers and larger numbers shrinks
  + can make skewed data normally distributed

### Log transformation

-   for more see Section 5.4 in @winter_statistics_2019

-   the R funtion `log()` computes the 'natural logarithm' (and is the inverse of the exponential `exp()`)
  -   `log()` makes large numbers smaller
  -   `exp()` makes small numbers larger

```{r}
#| output-location: column-fragment
log(0)
```
```{r}
#| output-location: column-fragment
log(1:10)
```
```{r}
#| output-location: column-fragment
log(c(10,20,30,40,100))
```

```{r}
#| output-location: column-fragment
exp(1:10)
```
```{r}
#| output-location: column-fragment
exp(log(1:10))
```

## Fit model (log-transformed)

-   continuous variables truncated at 0 typically have a *positive skew*
    -   a lot of small values (e.g., `tt` \< 500ms), with some larger values (\> `tt` 1000)
    -   this usually means our residuals are also positively skewed, i.e., not normally distributed
-   so we typically log-transform raw reading/reaction times for our linear models

```{r}
#| output-location: column-fragment
# fit simple linear model with log
fit_fp_lifetime_log <- df_crit_verb %>%
  filter(fp > 0) %>% # important! you can't log transform 0
  lm(log(fp) ~ lifetime, data = .)
summary(fit_fp_lifetime_log)
```

### Check assumptions

:::: columns
::: {.column width="50%"}
Raw first-pass reading times
```{r}
plot(density(resid(fit_fp_lifetime)))
```
:::

::: {.column width="50%"}
Log first-pass reading times
```{r}
plot(density(resid(fit_fp_lifetime_log)))
```
:::
::::

### Check assumptions {-}

:::: columns
::: {.column width="50%"}
Raw first-pass reading times
```{r}
qqnorm(residuals(fit_fp_lifetime))
qqline(residuals(fit_fp_lifetime), col="red")
```
:::

::: {.column width="50%"}
Log first-pass reading times
```{r}
qqnorm(residuals(fit_fp_lifetime_log))
qqline(residuals(fit_fp_lifetime_log), col="red")
```
:::
::::

# Communicating your results

-   model summaries can be provided via tables and/or figures
    -   you should always report the t-values and p-values of an effect

```{r}
df_crit_verb |> 
  filter(fp > 0) |> 
  mutate(log_ff = log(fp)) |> 
  mutate(half = if_else(trial >= 104, "1st","2nd")) |> 
  ggpubr::ggqqplot(x = "log_ff")
```

```{r}
df_crit_verb |> 
  filter(fp > 0) |> 
  mutate(log_ff = log(fp)) |> 
  mutate(half = if_else(trial >= 104, "1st","2nd")) |> 
  ggpubr::ggqqplot( x = "log_ff",
                    color = "half")
```

```{r}
df_crit_verb |> 
  filter(fp > 0) |> 
  mutate(log_ff = log(fp)) |> 
  mutate(half = if_else(trial >= 104, "1st","2nd")) |> 
  ggpubr::ggqqplot( x = "log_ff",
                    color = "half",
                    facet.by = "px")
```

```{r}
plot(density(resid(fit_fp_lifetime)))
```

# Summary

-   we looked at some assumptions of linear models and how to check them

-   next, we'll look at a case with more than one predictor: **multiple** regression