---
title: "Reproducible analysis reports with eye-tracking reading time data"
subtitle: "SoSe2023"
author: "Daniela Palleschi"
institute: Humboldt-Universität zu Berlin
lang: en
date: 2023-04-12
format: 
  pdf:
    output-file: pdf-welcome.pdf
    toc: true
    number-sections: false
    colorlinks: true
    code-overflow: wrap
  html:
    output-file: sheet-welcome.html
    number-sections: false
    toc: true
    code-overflow: wrap
    code-tools: true
    self-contained: true
  revealjs:
    output-file: slides-welcome.html
    code-overflow: wrap
    theme: [dark]
    width: 1600
    height: 900
    progress: true
    scrollable: true
    # smaller: true
    slide-number: c/t
    code-link: true
    # logo: logos/hu_logo.png
    # css: logo.css
    incremental: true
    # number-sections: true
    toc: true
    toc-depth: 2
    toc-title: 'Overview'
    navigation-mode: linear
    controls-layout: bottom-right
    fig-cap-location: top
    font-size: 0.6em
    slide-level: 4
    self-contained: true
    title-slide-attributes: 
      data-background-image: logos/logos.tif
      data-background-size: 15%
      data-background-position: 50% 92%
editor_options: 
  chunk_output_type: console
bibliography: references/references.bib
csl: references/apa.csl
---

```{r setup, eval = T, echo = F}
knitr::opts_chunk$set(echo = F, # print chunks?
                      eval = T, # run chunks?
                      error = F, # print errors?
                      warning = F, # print warnings?
                      message = F, # print messages?
                      cache = F # cache?; be careful with this!
                      )
```

```{r, eval = T, cache = F}
# Create references.json file based on the citations in this script
rbbt::bbt_update_bib("_welcome.qmd")
```




# Welcome!

-   the language of instruction is English
-   Block course:
    -   April 12-14 (10am-4pm)
    -   June 30th (2-6pm)
    -   July 1st (10am-4pm)
    
# Course Moodle

Course name: Reproducible analysis reports with eye-tracking reading time data (Blockseminar)
Enrolment Key: Rmatey

Most documents are available as slides, html, and PDF. Choose whichever you prefer (I suggest html).

## Course description

- develop skills and know-how
  + create reproducible **reports & presentations** of eye-tracking reading data
  + common measures in **eye-tracking reading**
  + importance of **reproducible workflow**
  + **communicate** findings

- hands-on exercises in RStudio with the R programming language
  + data **wrangling** (`tidyverse`)
  + data **visualisation** (`ggplot2`), 
  + descriptive and inferential **statistics** (`lme4` and `lmerTest`)
  
## Course credits

- 4 LP
  + attendance and participation: 1LP
  + In-class exercises and preparation: 1LP
  + Assignments: 2 LP
    1. Reproducible (pilot) analysis report + Pre-registration
    2. Reproducible analysis report

# Reading list

-   this course does not have a heavy reading load, but a few readings are strongly recommended:
    -   Open Science: @kathawalla_easing_2021 
    -   Eye-tracking reading: @clifton_eye_2007; @vasishth_what_2013; 
    -   A short recommendation for statistics for psycholinguists: @vasishth_statistical_2016
    -   Statistics for Linguistics (textbook): @winter_statistics_2019 (E-book available via Grimm)

## Further readings

- there are lots of useful resources out there, specifically:
  + Bodo Winter's tutorials on linear (mixed) models [@winter_linear_2013; @winter_very_2014]
  + the [PsyTeachR](https://psyteachr.github.io/) website is a *great* resource for hands-on stats and/or data analysis in R from the University of Glasgow School of Psychology and Neuroscience
  

# Session Info {.unlisted .unnumbered visibility="uncounted"}

Save your session info at the end of each document. Our results very often depend on the version of R/RStudio/a package we used. This is a great first step towards creating a reproducible workflow!

```{r}
sessionInfo()
```

# References {.unlisted .unnumbered visibility="uncounted"}

::: {#refs custom-style="Bibliography"}
:::
