---
title: "Reproducible analysis reports with eye-tracking reading time data"
subtitle: "Syllabus SoSe 2023"
author: "Daniela Palleschi"
institute: Humboldt-Universität zu Berlin
lang: en
date: 2023-04-12
format: 
  revealjs:
    output-file: slides-syllabus.html
    code-overflow: wrap
    theme: [dark]
    width: 1600
    height: 900
    progress: true
    scrollable: true
    # smaller: true
    slide-number: c/t
    code-link: true
    # logo: logos/hu_logo.png
    # css: logo.css
    incremental: true
    # number-sections: true
    toc: true
    toc-depth: 2
    toc-title: 'Overview'
    navigation-mode: linear
    controls-layout: bottom-right
    fig-cap-location: top
    font-size: 0.6em
    slide-level: 4
    chalkboard: true
    title-slide-attributes: 
      data-background-image: logos/logos.tif
      data-background-size: 15%
      data-background-position: 50% 92%
  pdf:
    output-file: pdf-syllabus.pdf
    toc: true
    number-sections: false
    colorlinks: true
    code-overflow: wrap
  html:
    output-file: sheet-syllabus.html
    number-sections: false
    toc: true
    code-overflow: wrap
editor_options: 
  chunk_output_type: console
bibliography: references/references.json
biblio-style: apalike
---

```{r setup, eval = T}
knitr::opts_chunk$set(echo = F, # print chunks?
                      eval = T, # run chunks?
                      error = F, # print errors?
                      warning = F, # print warnings?
                      message = F, # print messages?
                      cache = F # cache?; be careful with this!
                      )
```


```{r, cache = F}
# Create references.json file based on the citations in this script
rbbt::bbt_update_bib("syllabus_repro_sose23.qmd")
```

# Course description

This course will provide students with the skills and know-how needed to create reproducible reports and presentations of eye-tracking reading data. A brief introduction will be given into common measures in eye-tracking reading and the importance of developing a reproducible workflow, followed by hands-on exercises in RStudio with the R programming language. The main skills developed include data wrangling (with the tidyverse package), data visualisation (with the ggplot2 package), and running and communicating descriptive and inferential statistics. By the end of the course, students will be able to apply what they learned to a variety of data types in both academic and professional settings. This course is aimed at students who have some practical experience with R and RStudio, although this is not a strict requirement. Students who cannot bring their own laptop to class should contact the instructor as early as possible, so an alternative laptop can be organised. The language of instruction is English.

## Skills and know-how

At the end of the course, you will:

-   have an understanding of common eye-tracking reading measures
-   be able to visualise reading times
-   be able to run linear mixed models on reading times
-   be able to set-up and maintain a reproducible workflow in RStudio

# Course credits

- to earn credits for this course, you need to hand in:
  + 2 in-class quizzes (15 minutes each)
  + mid-course analysis report (due May 31st)
  + final analysis report (Due August 15th)

- for the first report, we will do some work in-class
  + e.g., data wrangling and tidying
  + independent work: turning it into a *reproducible report*
- for the second report, you will get new data
  + you will need to do the wrangling and analyses independently
  + and turn it into a *reproducible report*

## Important dates

- May 31st: first report due
- August 15th: second report due

## Schedule overview

+---------+-----------------------+------------------------------------------+
| Session | Date                  | Topic(s)                                 |
+=========+=======================+==========================================+
| 1       | Wed. April 12, 2023   | Eye-tracking during reading              |
|         |                       |                                          |
|         | 10:15 - 11:45         | Reading: @vasishth_what_2013             |
+---------+-----------------------+------------------------------------------+
| 2       | Wed. April 12, 2023   | Reproducible workflow with RStudio       |
|         |                       |                                          |
|         | 12:15 - 1:45          | Reading:                                 |
+---------+-----------------------+------------------------------------------+
| 3       | Wed. April 12, 2023   | Working with a dataset: data wrangling   |
|         |                       |                                          |
|         | 2:15 - 3:45           | Reading: @biondo_yesterday_2022          |
+---------+-----------------------+------------------------------------------+
| 4       | Thurs. April 13, 2023 | Data Wrangling and summary statistics    |
+---------+-----------------------+------------------------------------------+
| 5       | Thurs. April 13, 2023 | Data Visualisation                       |
+---------+-----------------------+------------------------------------------+
| 6       | Thurs. April 13, 2023 | Communicating your results               |
+---------+-----------------------+------------------------------------------+
| 7       | Fri. April 14, 2023   | Linear regression                        |
+---------+-----------------------+------------------------------------------+
| 8       | Fri. April 14, 2023   | Multiple regression                      |
+---------+-----------------------+------------------------------------------+
| 9       | Fri. April 14, 2023   | Linear mixed models                      |
+---------+-----------------------+------------------------------------------+
|         | ***April 15-June 29***| ***Individual work***                    |
+---------+-----------------------+------------------------------------------+
| 10      | Fri. June 30, 2023    | (Generalised) linear mixed models        |
+---------+-----------------------+------------------------------------------+
| 11      | Fri. June 30, 2023    | (Generalised) linear mixed models        |
+---------+-----------------------+------------------------------------------+
| 12      | Sat. July 1, 2023     |                                          |
+---------+-----------------------+------------------------------------------+
| 13      | Sat. July 1, 2023     |                                          |
+---------+-----------------------+------------------------------------------+
| 14      | Sat. July 1, 2023     |                                          |
+---------+-----------------------+------------------------------------------+

\newpage

# Reading list {-}

::: {#refs custom-style="Bibliography"}
:::

