---
title: "Setting up R and RStudio"
author: "Daniela Palleschi"
institute: Humboldt-Universität zu Berlin
lang: en
date: 2023-04-12
format:
  html:
    output-file: sheet-setup.html
    toc: true
    toc-depth: 2
    toc-title: 'Overview'
    code-link: true
    code-overflow: wrap
    code-tools: true
    number-sections: true
    self-contained: true
  pdf:
    output-file: pdf-setup.pdf
    toc: true
    number-sections: false
    colorlinks: true
    code-overflow: wrap
editor_options: 
  chunk_output_type: console
---

```{r, echo = F}
knitr::opts_chunk$set(eval = F, # evaluate = T for REPRODUCIBLE analyses
                      echo = T, # 'print code chunk?'
                      message = F, # print messages?
                      error = F, # render even if errors encountered?
                      warning = F) # print warnings?
```

This document outlines the steps required to set up our reproducible workflow. @sec-R provides an overview of installing R, RStudio, and required packages. These steps are required. @sec-tinytex describes how to install TinyTex, which is required to render LaTeX-style documents (i.e., PDFs) in R.

@sec-zotero provides tips on installing Zotero and related addins for integrations with R and RStudio. This is optional, and slightly more advanced.

# R and RStudio {#sec-R}

Download R [here](https://cran.r-project.org/).

Download RStudio [here](https://www.rstudio.com/products/rstudio/download/).

You must first download R, then RStudio.

## Packages {#sec-packages}

R packages available on the Comprehensive R Archive Network, commonly known as CRAN repository, can be easily installed using the command `isntall.packages("packageName")`. Some packages we will be needing are:

- `here` package
- `tidyverse` family of packages
  -  will automatically include packages we need, like `dplyr` and `ggplot2`
  
To download multiple packages at once, use the 'concatenate' function in r (`c()`) inside `install.packages()`:

```{r, eval = F}
install.packages(c("here", "tidyverse"))
```

## Developer packages (optional)

Not all packages are available on CRAN. You will need to install them directly through the developer. Two packages that are commonly used to do this are `remotes` and `devools`. Let's try installing `remotes` and then installing the `rbbt`, an addin for Zotero (discussed below).

```{r}
if(!requireNamespace("remotes", quietly = TRUE)) install.packages("remotes")
```

```{r}
remotes::install_github("paleolimbot/rbbt")
```

## RStudio Global options (optional)

Here are my preferred global options (RStudio > Tools > Global options). I highly recommend following the settings for 'workspace' and 'R Sessions' to ensure reproducibility. The other settings you can play around with to find what you like.

  - General > Basic
    - **Workspace** (for reproducibile workflow!!!)
        + uncheck 'Restore .RData into workshapce at startup'
        + Save workspace to .RData on exit: ***Never***
    - **R Sessions**
        + uncheck 'Restore previously open source documents at startup'
  - Code > Display
    - General
      + Show white space characters
      + Allow scroll past end of document
      + highlight selected line
  - Appearance
    - Editor theme: Cobalt

# tinyTex {#sec-tinytex}

To render PDF documents with LaTeX under the hood, we need to install [tinytex](https://yihui.org/tinytex/). There are different ways to do this:

- run the following in the *terminal*: `quarto install tinytex`
- or in the console: `tinytex::install_tinytex()`

# Zotero (optional) {#sec-zotero}

Zotero is a very useful PDF reader/annotator tool that is helpful even if not integrated with R/RStudio. I suggest downloading it and reading all your PDFs in Zotero in the future. You can download Zotero here: https://www.zotero.org/download/

## `rbbts`: Citations with Zotero

There is a Zotero addin for RStudio that allows you to directly read-in your citations. To use the `rbbt` package installed above, we need to have Zotero installed (and Better BibTex).

```{r}
# Install rbbt Addin from GitHub to use Zotero
remotes::install_github("paleolimbot/rbbt")
```

To set-up a citation entry shortcut: RStudio > Tools > Modify Keyboard Shortcuts > enter 'Zotero' in the search > choose 'Insert Zotero Citations' > add a shortcut (I like Ctrl+K)


### Better BibTex

Install from the [website](https://retorque.re/zotero-better-bibtex/installation/).

- then in Zotero: Tools > Add-ins > Settings wheel > choose downloaded file > Restart Zotero

- set your citation keys: Zotero > Settings > Better BibTex > Citation key formula > (I like "zotero.clean")
  + if you need to update the BibTex key in your docs: 
      + highlight all your files in the Zotero Library, right-click > Better BibTex > Refresh BibTex key

