---
title: "Working with eye-tracking reading data in R"
subtitle: "Loading and eye-balling a dataset"
author: "Daniela Palleschi"
institute: Humboldt-Universität zu Berlin
lang: en
date: 2023-04-12
format:
  revealjs: 
  # all docs:
    output-file: slide-et_dataset.html
  # revealjs and html:
    # toc
    # code
    toc: true
    toc-depth: 2
    toc-title: 'Overview'
    code-link: true
    code-overflow: wrap
    code-tools: true
    css: styles.css
  # revealjs specific:
    theme: [dark]
    width: 1600
    height: 900
    progress: true
    font-size: 0.6em
    slide-level: 4
    slide-number: c/t
    navigation-mode: linear
    # smaller: true
    incremental: true
    scrollable: true
    controls-layout: bottom-right
    fig-cap-location: top
    title-slide-attributes: 
      data-background-image: logos/logos.tif
      data-background-size: 15%
      data-background-position: 50% 92%
    # to add drawings:
    # chalkboard:
    #   src: drawings/day1-drawings.json
    embed-resources: true
    # logo: logos/hu_logo.png
    # css: logo.css
    # number-sections: true
  html:
    output-file: sheet-et_dataset.html
    embed-resources: true
    toc: true
    toc-depth: 2
    toc-title: 'Overview'
    code-link: true
    code-overflow: wrap
    code-tools: true
    number-sections: false
  pdf:
    output-file: pdf-et_dataset.pdf
    toc: true
    number-sections: false
    colorlinks: true
    code-overflow: wrap
editor_options: 
  chunk_output_type: console
bibliography: references/references.json
csl: references/apa.csl
---

# Set-up {.unlisted .unnumbered visibility="uncounted"}

```{r}
knitr::opts_chunk$set(eval = T, # evaluate = T for REPRODUCIBLE analyses
                      echo = T, # 'print code chunk?'
                      message = F, # print messages?
                      error = T, # render even if errors encountered?
                      warning = F) # print warnings?
```

```{r}
#| code-line-numbers: "1,2"
library(here) # relative path
library(tidyverse) # tidy/transform
library(beepr) # beeps when code runs or fails
library(rbbt) # zotero plugin
```

```{r}
#| code-fold: true
#| code-summary: beep code
## play sound if error encountered
### from: https://sejohnston.com/2015/02/24/make-r-beep-when-r-markdown-finishes-or-when-it-fails/
options(error = function(){    # Beep on error
  beepr::beep(sound = "wilhelm")
  Sys.sleep(2) # 
  }
 )
## and when knitting is complete
.Last <- function() {          # Beep on exiting session
  beepr::beep(sound = "ping")
  Sys.sleep(6) # allow to play for 6 seconds
  }
```

```{r, eval = T, cache = F}
#| code-fold: true
#| code-summary: rbbt code
#| 
# Create references.json file based on the citations in this script:
# 1. make sure you have 'bibliography: references/references.json' in the YAML
# 2. create a new folder called 'references'
# 3. run:
rbbt::bbt_update_bib("_et_dataset.qmd")
```

# The Perfect Lifetime Effect {.smaller}

-   the English Present Perfect (e.g., *has done*) [e.g., @comrie_aspect_1976]
    -   must be used in temporal contexts that *include the present*
        -   *I have been sick* [***since last week***]{.underline}
        -   \**I have been sick* [***last year***]{.underline}
-   The Lifetime Effect
    -   a referent's lifetime (dead/alive) constrains verb tense in certain circumstances [e.g., @mittwoch_tenses_2008]
        -   \**Queen Elizabeth II* [***is***]{.underline} *the British monarch.*
        -   \**King Charles III* [***was***]{.underline} *the British monarch.*
-   the Perfect Lifetime Effect
    -   the (English) Present Perfect cannot be used to describe events of a dead person [e.g., @mittwoch_english_2008]
        -   \**Queen Elizabeth II* [***has met***]{.underline} *many politicians.*
        -   *King Charles III* [***has met***]{.underline} *many politicians.*

## Our first dataset

::: columns
::: {.column width="40%"}
-   referent-lifetime context
    -   dead/alive
-   critical sentence
    -   Present Perfect/Simple Future
-   binary naturalness judgement to end trial
    -   accept/reject
:::

::: {.content-visible when-format="revealjs" when-format="html"}
::: {.column width="60%"}
![](media/livingPP_pilot1.mp4){fig-align="center" width="100%"}
:::
:::
:::

## Design description

::: columns
::: {.column width="50%"}
-   2x2 mixed design
    -   two 2-level factors (2x2 = 2-level x 2-level)
        -   factor 1: lifetime (levels: dead, alive)
        -   factor 2: tense (levels: PP, SF)
:::

::: {.column width="50%"}
|     | alive                     | dead                    |
|-----|---------------------------|-------------------------|
| PP  | Eddie Redmayne...has won  | Gene Kelly...\*has won  |
| SF  | Eddie Redmayne...will win | Gene Kelly...\*will win |
:::
:::

::: columns
::: {.column width="50%"}
-   predictors/independent variables
    -   lifetime
    -   tense
:::

::: {.column width="50%"}
-   measure/dependent variables (verb region)
    -   first-fixation time (milliseconds)
    -   first-pass reading time (ms)
    -   regression path duration (ms)
    -   total reading time (ms)
:::
:::

### Repeated measures design

-   observations are repeated e.g., multiple data points per participant, and per item across participants
    -   essentially, data are not independent
    -   e.g., each participant will have their own reading speed, some items might be systematically less acceptable for some unforeseen reason, etc.

# Working with the data

::: columns
::: {.column width="30%"}
Day 1

1.  load the data
2.  inspect data
    -   eyeball data structure
    -   print summaries
    -   plot data distributions
:::

::: {.column width="30%"}
Day 2

3.  tidy data
4.  visualise data
5.  communicate data
:::

::: {.column width="30%"}
Day 3

6.  analyse data
    -   confirmatory (a priori)
    -   exploratory (post-hoc)
7.  report analyses
:::
:::

## Install packages

```{r, eval = F}
install.packages("tidyverse")
install.packages("here")
```

-   install
    -   only do once
    -   ...or when you working on a new computer
    -   ...or after updating R
-   might be a wise idea to create a script just for installing packages
    -   can save time/energy when updating R

## Load packages

```{r}
library(tidyverse)
library(here)
```

-   load packages
    -   needed at the start of each session

## Load dataset

```{r}
df_lifetime <- readr::read_csv(here::here("data/data_lifetime_pilot.csv"))
```

-   N.B., `readr::read_csv` can be read as "`read_csv()` function in the `readr` package"
    -   i.e., `package::function()`
    -   you only need to use this syntax if you haven't loaded the specific package yet (maybe because you only need it once), or if a function name is included in multiple packages (i.e., there's a discrepancy in what `read_csv` could be referring to)
    -   why did I use it here?

## {.unlisted}
::: {.callout-tip collapse="true"}

### `here` package

Using the `here` package, we can access files *relative* to where our .RProj is stored.

In 'olden times', we had to specify the file path with something like:

```{r, eval = F}
# load in data from an *absolute* file path
df_lifetime <- read_csv("Users/yournamehere/Documents/SoSe2023/ET_reading/data/data_lifetime_pilot.csv")
```

Or, we'd set an *absolute* path as our working directory, to which all other file paths were *relative*

```{r, eval = F}
# set *absolute* path as working directory
setwd("Users/username/Documents/SoSe2023/ET_reading")

# load in data *relative* to our wd
df_lifetime <- read_csv("data/data_lifetime_pilot.csv")
```

This meant that if I sent my project folder to somebody else, they wouldn't be able to run my code because they would have to change the *absolute* file path to match their machine.
:::

## Inspect dataset

-   there are several different things you can inspect
    -   and different ways to accomplish those things
-   the first thing I usually do is look at the column/variable names

### `names()` {.smaller}

:::: {columns}

::: {.column width="60%"}
-   the names in all caps are variables created during the experiment
    -   i.e., they are our recorded *data*, mainly what we wanted to measure: dependent variables (DV)
    -   also includes some information about the experiment set-up per participant
-   the other names are variables from my stimuli lists
    -   i.e., they mostly contain our independent variables (IV)/stimuli
-   we typically want to see what effect our IVs had on any given DVs
-   variable descriptions can be found on the Moodle: Data > Documentation
:::

::: {.column width="40%"}
```{r}
names(df_lifetime)
```

:::

::::





### `rename()`

-   the dependent variable names are pretty clunky, let's rename a few:
    -   `RECORDING_SESSION_LABEL` corresponds to a single participant
    -   `TRIAL_INDEX` logged the trial number
    -   `EYE_USED` logged which eye was tracked

```{r}
#| code-fold: true
df_lifetime <- df_lifetime %>%
  rename("px" = RECORDING_SESSION_LABEL,
         "trial" = TRIAL_INDEX,
         "eye" = EYE_USED)
```

#### Naming variables

::: callout-tip
#### Naming conventions

It's wise to keep variable and object names concise but informative

-   all lowercase means fewer key strokes overall
-   separate words with either periods or underscores, e.g., `trial.index` or `trial_index`
-   e.g., we called our dataset `df_lifetime` because it is a dataframe (`df`) with data from our lifetime experiment
:::

### Data structure

-   datasets typically contain a lot of rows and columns
    -   so we want to get a feel for how the data is structured

::: panel-tabset
#### with base R

```{r}
#| output-location: fragment
head(df_lifetime)
```

#### with the tidyverse pipe

```{r}
#| code-line-numbers: "|1|2"
df_lifetime %>%
  head()
```

#### with the native R pipe (Ctrl/Cmd+Shift+M)

```{r}
#| code-line-numbers: "|1|2"

df_lifetime |> 
  head()
```
:::

### `head()` function

-   *prints* the first 6 rows of your data
    -   you can also specify the number of rows

```{r}
#| code-fold: true
#| output-location: fragment
df_lifetime %>%
  head(n = 2)
```

### `head()` function task {.unlisted .unnumbered visibility="uncounted"}

::: {.callout-tip collapse="true"}
#### Exercise: `head()`

1.  print only 2 rows using whichever syntax you prefer
2.  change `n = 2` to some other number and print
3.  run `?head` in the `Console`
    -  find the opposite function (i.e., prints last rows) in the function description?
4.   run this function with `df_lifetime` as argument; how many rows does it print as default?
5.   play with `n =` in this function to print some other number of rows
:::

### `tail()` function

-   prints the last rows of a dataframe (or matrix, vector, table, or function)

```{r}
df_lifetime %>%
  tail()
```

### `names()`

-   prints the column/variable names

```{r}
df_lifetime %>%
  names()
```

### `summary()`

-   prints a summary of each variable (column)

```{r}
df_lifetime %>%
  summary()
```

### Exercise

Take some time to explore the dataset.

-   double click on the dataset name in the Environment pane to view it like a spreadsheet
-   look at the names, can you figure out what they represent?

## class types

-   there are difference classes of data that R can read
    -   the function `class()` takes as its argument an object or number

```{r}
#| code-line-numbers: "|1|2"
#| output-location: column-fragment
df_lifetime$rt %>%
  class()
```

## {.unlisted}

::: {.callout-tip collapse="true"}
## Selecting a column

```{r}
# with column index
df_lifetime[2] %>% summary()
```

```{r}
# with column name
df_lifetime[,"trial"] %>% summary()
```

```{r}
# with data$column_name
df_lifetime$trial %>% summary()
```

```{r}
# with the tidyverse: select()
df_lifetime %>% 
  select(trial) %>%
  summary()
```
:::

### `character` class

-   contain *strings*: collection of characters (i.e., text)
-   there's no grouping in character variables
    -   each value is considered 'unique' and assumed to not be repeated
-   we usually aren't interested in character class variables
    -   unless e.g., we have unique values per row (e.g., if a participant gave a free-text answer)
    -   or perhaps we have stored some stimuli sentences
        -   although this would arguably be better as a 'category', since there should be multiple trials across participants that contain the same sentences

### `numeric` class

-   variables with numeric values, usually some variable we'd want to compute summaries on, e.g., means
-   sometimes we don't want numbers to be stored as numeric class, however
    -   this is the case for our variables `yes_press` and `KeyPress` (with `4` or `5`)
-   the same is true for our variable `item_id`, which ranges from 1:120
    -   the numbers are just unique codes for our stimuli, the difference between `item 1` and `item 2` has nothing to do with the difference between the numbers `1` and `2`

### `factor` class

-   we typically want *grouping* variables to be `factor` class
    -   factors contain ***categorical*** data
    -   any number that could be replaced with some other label should be a factor
-   region of interest (ROI) = 1:7
    -   but we want to know how many observations per region, the number is not informative
    -   ROI could alternatively be coded as, e.g., "adverb", "pronoun", "verb", "spillover"

### `factor` class {.unlisted .unnumbered visibility="uncounted"}

-   let's change `df_lifetime$yes_press` to `factor`
    -   using the `mutate()` verb from `dplyr`
    -   and `as_factor()` from `forcats`

```{r}
#| cold-fold: true
#| code-line-numbers: "|"
#| output-location: fragment
# change yes_press to factor
df_lifetime %>%
  mutate(yes_press = as_factor(yes_press))
```

### multiple arguments in a verb

-   we can also change multiple columns at once:

```{r}
# change ROI & label to factor
df_lifetime %>%
  mutate(KeyPress = as_factor(KeyPress),
         item_id = as_factor(item_id))
```

### Pop quiz

1.  Which class *should* the following variables be (`numeric`, `factor`, or `character`)?:

    -   participant ID
    -   trial number
    -   first-pass reading time
    -   regression path duration
    -   regressions in
    -   context sentence
    -   lifetime
    -   tense
    -   celebrity name

2.  change them to these class types, and print a summary

3.  save and render the document


## Plot the data


::: {.column width="100%"}
-   at this stage we want to explore the data
    -   distribution
        -   peaks, spread
    -   boundaries
:::


:::: columns
::: {.column width="33%"}
Histogram

```{r}
#| output-location: fragment
hist(df_lifetime$IA_FIRST_RUN_DWELL_TIME)
```
:::

::: {.column width="33%"}
Boxplot

```{r}
#| output-location: fragment
boxplot(df_lifetime$IA_FIRST_RUN_DWELL_TIME)
```
:::

::: {.column width="33%"}
Scatterplot

```{r}
#| output-location: fragment
plot(df_lifetime$IA_FIRST_RUN_DWELL_TIME)
```
:::

::::

### Plotting two variables

Scatterplot

```{r}
#| output-location: fragment
plot(df_lifetime$IA_FIRST_FIXATION_DURATION, df_lifetime$IA_FIRST_RUN_DWELL_TIME)
```

### Exercise

In your Quarto document:

1.  create a heading 'Data exploration' 
  + briefly describe the data
2.  For each of our depenent variables: 
  + create a subheading
  + calculate the mean and standard deviation of the variable (`mean()`, `sd()`) + create a boxplot of the variable
3.  Render the document often to make sure it runs
4.  Upload the source file (day1-nachname_vorname.qmd) to Moodle
5.  download the source file below yours in the list to the same folder, and try to run it
  +   does it run?

### {.unlisted}

::: callout-tip
#### print options

-   each code chunk can have different print options:
    -   `eval = FALSE`: do not evaluate this chunk
    -   `include = FALSE` evaluate this chunk but don't show it or its results
    -   `echo = FALSE` print this chunk code
    -   `message = FALSE`/`warning = false` don't print warnings or messages
    -   `error = TRUE` continue rendering document even if there's an error
        -   do not use `error = TRUE` for final versions! You want to make sure things work as they should

````markdown
`r ''````{r, eval = T, echo = T, results = "asis", warning}
code here
```
````

or
````markdown
`r ''````{r}
#| eval: false
code here
```
````
:::

# Session Info {.unlisted .unnumbered visibility="uncounted"}

```{r}
#| code-fold: true
#| code-summary: "Show Session Info"
sessionInfo()
```

# References {.unlisted .unnumbered visibility="uncounted"}

::: {#refs custom-style="Bibliography"}
:::
