---
title: "Working with eye-tracking reading data in R"
subtitle: "Biondo et al. (2021)"
author: "Daniela Palleschi"
institute: Humboldt-Universität zu Berlin
lang: en
date: 2023-04-12
format:
  revealjs: 
  # all docs:
    output-file: slide-et_dataset.html
  # revealjs and html:
    # toc
    # code
    toc: true
    toc-depth: 2
    toc-title: 'Overview'
    code-link: true
    code-overflow: wrap
    code-tools: true
    css: styles.css
  # revealjs specific:
    theme: [dark]
    width: 1600
    height: 900
    progress: true
    font-size: 0.6em
    slide-level: 4
    slide-number: c/t
    navigation-mode: linear
    # smaller: true
    incremental: true
    scrollable: true
    controls-layout: bottom-right
    fig-cap-location: top
    title-slide-attributes: 
      data-background-image: logos/logos.tif
      data-background-size: 15%
      data-background-position: 50% 92%
    # to add drawings:
    chalkboard:
      src: drawings/day1-drawings.json
    # logo: logos/hu_logo.png
    # css: logo.css
    # number-sections: true
  html:
    output-file: sheet-et_dataset.html
    toc: true
    toc-depth: 2
    toc-title: 'Overview'
    code-link: true
    code-overflow: wrap
    code-tools: true
    number-sections: false
  pdf:
    output-file: pdf-et_dataset.pdf
    toc: true
    number-sections: false
    colorlinks: true
    code-overflow: wrap
editor_options: 
  chunk_output_type: console
bibliography: references/references.json
biblio-style: apalike
---

# Set-up {.unlisted .unnumbered visibility="uncounted"}

```{r}
knitr::opts_chunk$set(eval = T, # evaluate = T for REPRODUCIBLE analyses
                      echo = T, # 'print code chunk?'
                      message = F, # print messages?
                      error = T, # render even if errors encountered?
                      warning = F) # print warnings?
```

```{r}
library(rbbt) # zotero plugin
library(here) # relative path
library(tidyverse) # tidy/transform
library(beepr) # beeps when code runs or fails
```

```{r}
## play sound if error encountered
### from: https://sejohnston.com/2015/02/24/make-r-beep-when-r-markdown-finishes-or-when-it-fails/
options(error = function(){    # Beep on error
  beepr::beep(sound = "wilhelm")
  Sys.sleep(2) # 
  }
 )
## and when knitting is complete
.Last <- function() {          # Beep on exiting session
  beepr::beep(sound = "ping")
  Sys.sleep(6) # allow to play for 6 seconds
  }
```

```{r, eval = T, cache = F}
# Create references.json file based on the citations in this script:
# 1. make sure you have 'bibliography: references/references.json' in the YAML
# 2. create a new folder called 'references'
# 3. run:
rbbt::bbt_update_bib("_et_dataset.qmd")
```

# The Perfect Lifetime Effect {.smaller}

-   the English Present Perfect (e.g., *has done*) [e.g., @comrie_aspect_1976]
    -   must be used in temporal contexts that *include the present*
        -   *I have been sick* [***since last week***]{.underline}
        -   \**I have been sick* [***last year***]{.underline}
-   The Lifetime Effect
    -   a referent's lifetime (dead/alive) constrains verb tense in certain circumstances [e.g., @mittwoch_tenses_2008]
        -   \**Queen Elizabeth II* [***is***]{.underline} *the British monarch.*
        -   \**King Charles III* [***was***]{.underline} *the British monarch.*
-   the Perfect Lifetime Effect
    -   the (English) Present Perfect cannot be used to describe events of a dead person [e.g., @mittwoch_english_2008]
        -   \**Queen Elizabeth II* [***has met***]{.underline} *many politicians.*
        -   *King Charles III* [***has met***]{.underline} *many politicians.*

## Our first dataset

-   eye-tracking during reading
    -   participants read a referent-lifetime context, followed by a critical sentence
        -   button-press when done reading
    -   critical sentences contained the Present Perfect or Simple Future
        -   button-press when done reading
-   each trial ended with a 7-point naturalness Likert scale

## Design description

:::: {.columns}

::: {.column width="100%"}

-   2x2 mixed design
    -   two 2-level factors (2x2 = 2-level x 2-level)
        -   factor 1: lifetime (levels: dead, alive)
        -   factor 2: tense (levels: PP, SF)
    - four conditions:
        - dead-PP, living-PP, dead-SF, living-SF
:::

::::

:::: {.columns}

::: {.column width="50%"}
-   predictors/independent variables
    -   lifetime
    -   tense
:::

::: {.column width="50%"}
-   measure/dependent variable(s)
    -   first-fixation time (milliseconds)
    -   first-pass reading time (ms)
    -   regression path duration (ms)
    -   tt (ms)
:::

::::

## Repeated measures design

- repeated measures design
  - observations are repeated e.g., multiple data points per participant, and per item within and across participants
  - essentially, data are not independent

- relevant terms
  - counterbalancing: each condition presented equally across participants, across experimental lists, 
  - Latin-square design

# Working with the data

::: columns
::: {.column width="30%"}
Day 1

1.  load the data
2.  inspect data
    -   eyeball data structure
    -   print summaries
    -   plot the data
:::

::: {.column width="30%"}
Day 2

3.  tidy data
4.  transform data
5.  visualise data
:::

::: {.column width="30%"}
Day 3

6.  analyse data
    -   confirmatory (a priori)
    -   exploratory (post-hoc)
7.  report analyses
:::
:::

## 1. Load/install packages

- install
  + only do once
  + ...or when you working on a new computer
  + ...or after updating R
- might be a wise idea to create a script just for installing packages
  + can save time/energy when updating R
  
```{r}
install.packages("tidyverse")
install.packages("here")
```

- load packages
  + needed at the start of each session
  
```{r}
library(tidyverse)
library(here)
```

## 2. Load dataset

```{r}
df_lifetime <- readr::read_csv(here::here("data/data_lifetime_pilot.csv"))
```

- N.B., `readr::read_csv` can be read as "`read_csv()` function in the `readr` package"
  + i.e., `package::function`
  + you only need to use this syntax if you haven't loaded the specific package yet (maybe because you only need it once), or if a function name is included in multiple packages (i.e., there's a discrepancy in what `read_csv` could be referring to)
  + why did I use it here?

::: {.callout-tip collapse="true"}
### `here` package

Using the `here` package, we can access files *relative* to where our .RProj is stored.

In 'olden times', we had to specify the file path with something like:

```{r, eval = F}
# load in data from an *absolute* file path
df_lifetime <- read_csv("Users/yournamehere/Documents/SoSe2023/ET_reading/data/data_lifetime_pilot.csv")
```

Or, we'd set an *absolute* path as our working directory, to which all other file paths were *relative*

```{r, eval = F}
# set *absolute* path as working directory
setwd("Users/username/Documents/SoSe2023/ET_reading")

# load in data *relative* to our wd
df_lifetime <- read_csv("data/data_lifetime_pilot.csv")
```

This meant that if I sent my project folder to somebody else, they wouldn't be able to run my code because they would have to change the *absolute* file path to match their machine.
:::

## 2. Inspect dataset

-   there are several different things you can inspect
    -   and different ways to accomplish those things

### Data structure

-   datasets typically contain a lot of *rows*
    -   so we want to get a feel for how the data is structured

::: panel-tabset
#### with base R

```{r}
#| output-location: fragment
head(df_lifetime)
```

#### with the tidyverse pipe

```{r}
#| code-line-numbers: "|1|2"

library(tidyverse)
df_lifetime %>%
  head()
```

#### with the native R pipe (Ctrl/Cmd+Shift+M)

```{r}
#| code-line-numbers: "|1|2"

df_lifetime |> 
  head()
```

::::

### `head()` function

-   *prints* the first 6 rows of your data
    -   you can also specify the number of rows

```{r}
#| code-fold: true
#| output-location: fragment
df_lifetime %>%
  head(n = 2)
```

### `head()` function task {.unlisted .unnumbered visibility="uncounted"}

::: {.callout-tip collapse="true"}
#### Task: `head()`

1.  print only 2 rows, but using the syntax for a single line of code
2.  change `n = 2` to some other number and print
3.  run `?head` in the `Console`

-   find the opposite function (i.e., prints last rows) in the function description?
-   run this function with `df_lifetime` as argument; how many rows does it print as default?
-   play with `n =` in this function to print some other number of rows
:::

### `tail()` function

-   prints the last rows of a dataframe (or matrix, vector, table, or function)

```{r}
df_lifetime %>%
  tail()
```

### `names()`

-   prints the column/variable names

```{r}
df_lifetime %>%
  names()
```

### `summary()`

-   prints a summary of each variable (column)

```{r}
df_lifetime %>%
  summary()
```

### Exercise

Take some time to explore the dataset.
  
  - double click on the dataset name in the Environment pane to view it like a spreadsheet
  - look at the names, can you figure out what they represent?

## class types

-   there are difference classes of data that R can read
    -   the function `class()` takes as its argument an object or number

```{r}
#| code-line-numbers: "|1|2"
df_lifetime$RT_value %>%
  class()
```

::: {.callout-tip collapse="true"}
# Selecting a column

```{r}
# with column index
df_lifetime[2] %>% summary()
```

```{r}
# with column name
df_lifetime[,"TRIAL_INDEX"] %>% summary()
```

```{r}
# with data$column_name
df_lifetime$TRIAL_INDEX %>% summary()
```

```{r}
# with the tidyverse: select()
df_lifetime %>% 
  select(TRIAL_INDEX) %>%
  summary()
```

:::

### `numeric` class

### `integer` class

### `factor` class

-   we typically want things *grouping* variables to be `factor` class
    -   factor class is ***categorical*** data
    -   any number that could be replaced and maintain its value should be a factor
-   region of interest (ROI) = 1:7
    -   but we want to know how many observations per region, the number is not informative
    -   ROI could alternatively be coded as, e.g., "adverb", "pronoun", "verb", "spillover"

### `factor` class {.unlisted .unnumbered visibility="uncounted"}

-   let's change `df_lifetime$ROI` to `factor`
    -   using `mutate()` from dplyr
    -   and `as_factor()` from forcats

```{r}
#| code-line-numbers: "|"
#| output-location: fragment
# change ROI & label to factor
df_lifetime %>%
  mutate(TRIAL_INDEX = as_factor(TRIAL_INDEX)) %>%
  summary()
```

### `character` class

-   we usually aren't interested in character class variables
    -   unless e.g., we have unique values per row (e.g., if a participant gave a free-text answer)
    -   or perhaps we have stored some stimuli sentences
        -   although this would arguably be better as a factor

### Pop quiz

1.  of the classes `numeric`, `factor`, and `character`:
    -   participant ID
    -   trial number
    -   first-pass reading time
    -   regression path duration
    -   regressions in
    -   verb type
    -   adverb type

2.  change them to these class types, and print a summary

3.  save render the document

4.  upload the source file (day1-nachname_vorname.qmd) to XYZ

5.  download the source file below yours in the list to the same folder, and try to run it

-   does it run?

## 3. Plot the data

::: columns
::: {.column width="50%"}
-   at this stage we want to explore the data
    -   distribution
        -   peaks, spread
    -   boundaries
:::
::::

::: columns
::: {.column width="40%"}
Histogram

```{r}
#| output-location: fragment
hist(df_lifetime$IA_FIRST_RUN_DWELL_TIME)
```
:::

::: {.column width="40%"}
Boxplot

```{r}
#| output-location: fragment
boxplot(df_lifetime$IA_FIRST_RUN_DWELL_TIME)
```
:::

::: {.column width="40%"}
Scatterplot

```{r}
#| output-location: fragment
plot(df_lifetime$IA_FIRST_RUN_DWELL_TIME)
```
:::

::: {.column width="40%"}
Scatterplot

```{r}
#| output-location: fragment
plot(df_lifetime$IA_FIRST_FIXATION_DURATION, df_lifetime$IA_FIRST_RUN_DWELL_TIME)
```
:::
:::::

### Exercise

In your Quarto document:

  1. create a heading 'Data exploration'
    + briefly describe the data
  2. For each of our depenent variables:
    + create a subheading
    + calculate the mean and standard deviation of the variable (`mean()`, `sd()`)
    + create a boxplot of the variable
  3. Render the document often to make sure it runs
  
::: callout-tip
#### print options

- each code chunk can have different print options:
  + `eval = FALSE`: do not evaluate this chunk
  + `include = FALSE` evaluate this chunk but don't show it or its results
  + `echo = FALSE` print this chunk code
  + `message = FALSE`/`warning = false` don't print warnings or messages
  + `error = TRUE` continue rendering document even if there's an error
    + do not use `error = TRUE` for final versions! You want to make sure things work as they should
    
`````markdown
`r ''````{r, eval = T, echo = T, results = "asis", warning}
```
or
`r ''````{r}
#| eval: false
```
`````

:::

# Session Info {.unlisted .unnumbered visibility="uncounted"}

```{r}
#| code-fold: true
sessionInfo()
```

# References {.unlisted .unnumbered visibility="uncounted"}

::: {#refs custom-style="Bibliography"}
:::
