---
title: "Data Visualisation with `ggplot2`"
subtitle: "Communicating your data"
author: "Daniela Palleschi"
institute: Humboldt-Universität zu Berlin
footer: "Data Visualisation"
lang: en
date: 2023-04-13
format:
  revealjs: 
    output-file: slides-data_viz.html
    theme: [dark]
    width: 1600
    height: 900
    progress: true
    # smaller: true
    scrollable: true
    slide-number: c/t
    code-link: true
    code-overflow: wrap
    code-tools: true
    # logo: logos/hu_logo.png
    # css: logo.css
    incremental: true
    # number-sections: true
    toc: true
    toc-depth: 1
    toc-title: 'Topics'
    navigation-mode: linear
    controls-layout: bottom-right
    fig-cap-location: top
    font-size: 0.6em
    slide-level: 4
    chalkboard: true
    title-slide-attributes: 
      data-background-image: logos/logos.tif
      data-background-size: 15%
      data-background-position: 50% 92%
  html:
    self-contained: true
    output-file: sheet-data_viz.html
    number-sections: true
    toc: true
    code-overflow: wrap
    code-tools: true
  pdf:
    output-file: pdf-data_viz.pdf
    toc: true
    number-sections: false
    colorlinks: true
    code-overflow: wrap
editor_options: 
  chunk_output_type: console
bibliography: references/references.json
csl: references/apa.csl
---

```{r}
knitr::opts_chunk$set(eval = T, # evaluate chunks
                      echo = T, # 'print code chunk?'
                      message = F, # 'print messages (e.g., warnings)?'
                      error = F, # stop when error encountered
                      warning = F) # don't print warnings
```

```{r}
## play sound if error encountered
### from: https://sejohnston.com/2015/02/24/make-r-beep-when-r-markdown-finishes-or-when-it-fails/
options(error = function(){    # Beep on error
  beepr::beep(sound = "wilhelm")
  Sys.sleep(2) # 
  }
 )
## and when knitting is complete
.Last <- function() {          # Beep on exiting session
  beepr::beep(sound = "ping")
  Sys.sleep(6) # allow to play for 6 seconds
  }
```

```{r, eval = T, cache = F}
# Create references.json file based on the citations in this script
# make sure you have 'bibliography: references.json' in the YAML
rbbt::bbt_update_bib("_data_viz.qmd")
```

# Data communication

## Load packages and data

```{r}
# load tidyverse
library(tidyverse)

# load data
df_lifetime <- readr::read_csv(here::here("data/tidy_data_lifetime_pilot.csv"), 
                               # for special characters
                               locale = readr::locale(encoding = "latin1") 
                               ) |>
  mutate_if(is.character,as.factor) |> # all character variables as factor
  filter(type == "critical", # only critical trials
         px != "px3") # this participant had lots of 0's for some reason
```

## Tables

- we can create summaries of our data

```{r, eval = F, echo = F}
# with Rmisc::summarySEwithin
df_lifetime |> 
  filter(region == "verb",
         !is.na(ff)) |> 
  Rmisc::summarySEwithin(measurevar="ff", withinvars ="condition")
```

```{r}
#| code-fold: true
# compute summary 
summary_ff <- df_lifetime |> 
  filter(region=="verb") |> 
  group_by(condition,lifetime,tense) %>%
  summarise(N = n(),
            mean.ff = mean(ff, na.rm = T),
            sd = sd(ff, na.rm = T)) %>%
  # compute standard error, confidence intervals, and lower/upper ci bounds
  mutate(se = sd / sqrt(N),
         ci = qt(1 - (0.05 / 2), N - 1) * se,
         lower.ci = mean.ff - qt(1 - (0.05 / 2), N - 1) * se,
         upper.ci = mean.ff + qt(1 - (0.05 / 2), N - 1) * se)
```

- and print the output with the `kable()` function from the `knitr` package
  + for extra customisation you can also use the `kableExtra` package (e.g., with the `kable_styling()` function)

```{r}
#| output-location: fragment

# install.packages("knitr") # if not yet installed
knitr::kable(summary_ff, digits=1,
             caption = "Table with summmary statistics for first-fixation duration at the verb region")
```

### Exercise

1. install the `knitr` package (`install.packages("knitr")`)
2. create an object with some summary statistics of the variable `rt`
  + call it `summary_rt`
3. use `kable()` from `knitr` to print a table

```{r}
#| echo: false
summary_rt <- df_lifetime |> 
  distinct(px,trial,.keep_all=T) |> 
  group_by(lifetime,tense,condition) %>%
  summarise(N = n(),
            mean.rt = mean(rt, na.rm = T),
            sd = sd(rt, na.rm = T))
```

```{r}
#| output-location: fragment
knitr::kable(summary_rt, digits=1,
             caption = "Summary of reaction times (ms) per condition")
```

# Plotting reading times

- reading times are (usually) continuous variables
  + as are e.g., reaction times
- they are truncated at 0, meaning they cannot have negative values
  + because of this, they tend to have a *skewed distribution*

# Plots with `ggplot2`

- `ggplot2` is part of the tidyverse (like `dplyr`)
  + uses a *layered grammar of graphics*
  + i.e., we build layers

# An example: histogram

```{r}
#| output-location: column-fragment
iris |> 
  ggplot(aes(x=Sepal.Length)) +
  geom_histogram() +
  labs(title = "A histogram",
       x = "variable")
```

## Start layering

```{r}
#| output-location: column
df_lifetime |> ggplot(aes(ff)) # aes = 'aesthetic'
```

## Add labels

```{r}
#| output-location: column
df_lifetime |> ggplot(aes(ff)) + 
  labs(title = "Histogram of first fixation times",
       x = "First fixation times (ms)")
```

## Add 

```{r, fig.cap="Distribution of first fixation times at the verb region (raw milliseconds)"}
#| output-location: column
df_lifetime |> ggplot(aes(ff)) + 
  labs(title = "Histogram of first fixataion times",
       x = "First fixation times (ms)") +
  geom_histogram()
```

## Add condition

```{r, fig.cap="Distribution of first fixation times at the verb region (raw milliseconds)"}
#| output-location: column
df_lifetime |> ggplot(aes(ff, fill = condition)) + 
  labs(title = "First fixataion times at the verb region",
       x = "First fixation times (ms)") +
  geom_histogram()
```

::: {.notes}
The colour here is STACKED!! i.e., not layered. Notice the distribution doesn't change from all grey to coloured
:::

## Customisation

- we can add arguments to our geoms
  + e.g., transparency: `alpha = ` takes a value between 0 to 1
- we can use `theme()` to customise font sizes, legend placement, etc.
- tehre are also popular preset themes, such as `theme_bw()` and `theme_minimal()`

::: panel-tabset
### `theme_bw()`
```{r, fig.cap="Distribution of first fixation times at the verb region (raw milliseconds)."}
#| output-location: column
#| code-line-numbers: "|4|5|"
df_lifetime |> ggplot(aes(ff, fill = condition)) + 
  labs(title = "Histogram of first fixataion times",
       x = "First fixation times (ms)") +
  geom_histogram(alpha=.5) +
  theme_bw()
```

### `theme_minimal()`
```{r, fig.cap="Distribution of first fixation times at the verb region (raw milliseconds)."}
#| output-location: column
#| code-line-numbers: "5"
df_lifetime |> ggplot(aes(ff, fill = condition)) + 
  labs(title = "Histogram of first fixataion times",
       x = "First fixation times (ms)") +
  geom_histogram(alpha=.5) +
  theme_minimal()
```
:::

# Distributions {.smaller}

- show the distribution of observations
  + so we can see where the data are clustered
  + and eyeball the shape of the distribution
- we already saw the histogram, which shows the number of observations per variable value
- **density plots** are another useful plot for visualising distributions

```{r}
#| output-location: column-fragment
iris |> 
  ggplot(aes(x=Sepal.Length)) +
  geom_density() +
  labs(title = "A density plot",
       x = "variable")
```

## Density plots

- below I just replaced `geom_histogram()` with `geom_density()`
  + I also filtered the data to include only values of `ff` above 0
- what is plotted along the y-axis? how does this differ from a histogram?

```{r, fig.cap="Distribution of first fixation times at the verb region (raw milliseconds)."}
#| code-line-numbers: "|2|6|"
#| output-location: column-fragment
df_lifetime |> 
  filter(ff > 0) |> 
  ggplot(aes(ff)) + 
  labs(title = "Histogram of first fixataion times",
       x = "First fixation times (ms)") +
  geom_density() +
  theme_minimal()
```

## Grouped density plots

- just like with histograms, we can look at the density plots of different subsets of the data with `aes(fill = )`
  + like region

```{r, fig.cap="Distribution of first fixation times at the verb region (raw milliseconds)."}
#| code-line-numbers: "|2|6|"
#| output-location: column-fragment
df_lifetime |> 
  filter(ff > 0) |> 
  ggplot(aes(ff, fill = region)) + 
  labs(title = "Histogram of first fixataion times",
       x = "First fixation times (ms)") +
  geom_density(alpha=.5) +
  theme_minimal()
```

### `facet_grid()`

- there are a lot of overlapping density curves, let's try to separate them with `facet_grid(x~y)`

```{r, fig.cap="Distribution of first fixation times at the verb region (raw milliseconds)."}
#| code-line-numbers: "|4|"
#| output-location: column-fragment
df_lifetime |> 
  filter(ff > 0) |> 
  ggplot(aes(ff, fill = region)) + 
  facet_grid(.~region) +
  labs(title = "Density plot of first fixataion times by region",
       x = "First fixation times (ms)") +
  geom_density(alpha=.5) +
  theme_bw()
```

- how would you describe the density plots of the different regions?

### re-ordering factors

- by default, factors will be ordered alphabetically
  + but we don't always want that
  + here, `verb-1` should be before `verb`
  
```{r}
#| output-location: column-fragment
df_lifetime <- df_lifetime %>%
  mutate(region = factor(region, 
                         levels = c("verb-1","verb","verb+1","verb+2","verb+3","verb+4")))

summary(df_lifetime$region)
```

#### Exercise


1. create a density plot with the fill colour set to `condition`, but:
  
- subset the data to only include the verb region
- you can decide if you want to use facets or to have the density curves overlayed
- your plot should look something like A or B:

```{r, echo = F}
#| output-location: fragment

# with facet
p_dens1 <- df_lifetime |>
  filter(ff > 0,
         region == "verb") |>
  ggplot(aes(ff, fill = condition)) +
  facet_grid(. ~ condition) +
  labs(title = "Density plot of first fixataion times (verb region)",
       x = "First fixation times (ms)",
       y = "Density",
       fill = "Condition") +
  geom_density(alpha = .5) +
  theme_bw() +
  theme(title = element_text(size=10),
        axis.text.x = element_text(angle = 45))

# without facet
p_dens2 <- df_lifetime |>
  filter(ff > 0,
         region == "verb") |>
  ggplot(aes(ff, fill = condition)) +
  # facet_grid(.~condition) +
  labs(title = "Density plot of first fixataion times
(verb region)",
       x = "First fixation times (ms)",
       y = "Density",
       fill = "Condition") +
  geom_density(alpha = .5) +
  theme_bw() +
  theme(plot.title = element_text(size=10))

# create legend

dens_legend <- cowplot::get_legend(p_dens1 + theme(legend.position = "bottom"))
dens_legend <- ggpubr::as_ggplot(dens_legend)

# combine
ggpubr::ggarrange(
ggpubr::ggarrange(p_dens1 + theme(legend.position = "none"),
                  p_dens2 +theme(legend.position = "none"),
                  labels = c("A","B"),
                  widths = c(.6,.4)),
dens_legend, nrow = 2, heights = c(.9,.1))

```


#### Extra exercise

2. Can you produce these plots?

```{r, echo = F}
p_tense_life <- df_lifetime |> 
  filter(ff > 0,
         region == "verb") |> 
  ggplot(aes(ff, fill = lifetime)) + 
  facet_grid(.~tense) +
  labs(title = "Density plot of first fixataion times (verb region)",
       x = "First fixation times (ms)") +
  geom_density(alpha=.5) +
  theme_bw() 

p_tense <- df_lifetime |> 
  filter(ff > 0,
         region == "verb") |> 
  ggplot(aes(ff, fill = tense)) + 
  # facet_grid(.~tense) +
  labs(title = "Density plot of first fixataion times (verb region)",
       x = "First fixation times (ms)") +
  geom_density(alpha=.5) +
  theme_bw() +
  theme(title = element_text(size=8))

p_lifetime <- df_lifetime |> 
  filter(ff > 0,
         region == "verb") |> 
  ggplot(aes(ff, fill = lifetime)) + 
  # facet_grid(.~tense) +
  labs(title = "Density plot of first fixataion times (verb region)",
       x = "First fixation times (ms)") +
  geom_density(alpha=.5) +
  theme_bw() +
  theme(title = element_text(size=8))

ggpubr::ggarrange(
  ggpubr::ggarrange(p_tense, p_lifetime,
                    labels = c("A","B")),
  ggpubr::ggarrange(NULL, p_tense_life,NULL,
                    nrow = 1,
                    widths = c(.1,.8,.1),
                    labels = c("", "C", "")),
  nrow = 2
)
```


## Scatterplots

- histograms and density plots plot a **single variable** along the x-axis
  + in most other plots the dependent (measure) variable is plotted along the y-axis by convention
- scatterplots plot the relationship between **two variables**

```{r}
#| output-location: column-fragment
iris |> 
  ggplot(aes(x=Sepal.Length, y=Sepal.Width)) +
  geom_point() +
  labs(title = "A scatterplot",
       x = "variable X",
       y = "variable Y")
```

### Scatterplots

- the figure below plots total reading times (verb region) to the verb region (x-axis) and reaction times to the critical sentence (y-axis)
  + what does each point represent?
  + how would you describe the relationship between the two variables?

```{r}
#| code-line-numbers: "4,7,8,9"
#| output-location: column-fragment
df_lifetime |>
  filter(ff > 0,
         region == "verb") |>
  ggplot(aes(x = tt, y = rt)) +
  labs(title = "Scatter plot of total reading times (verb region)
and reaction times (critical sentence)",
       x = "Total reading time (ms)",
       y = "Reaction time (ms)") +
  geom_point(alpha = .2) +
  theme_bw()
```

#### Exercise

1. Generate a scatterplot of total reading times and reaction times, with:
    - colour and shape set to condition
    + tip: these both belong in `aes()`
2. What information does this plot suggest?

```{r, eval = F, echo = F}
#| code-line-numbers: "4,7,8,9"
#| output-location: fragment
df_lifetime |>
  filter(ff > 0,
         region == "verb") |>
  ggplot(aes(x = tt, y = rt, colour = condition, shape = condition)) +
  labs(title = "Scatter plot of total reading times (verb region)
and reaction times (critical sentence)",
       x = "Total reading time (ms)",
       y = "Reaction time (ms)",
colour = "Condition", shape = "Condition") +
  geom_point(alpha = .5, size = 2) +
  theme_bw()
```

# Summary statistics

- measures of location: mean, median, mode
- measures of spread: (interquartile) range, standard deviation

::: {.content-visible when-format="revealjs"}
::: notes
IQR: middle 50% of data
:::
:::

## Boxplots

- boxplots provide information about the distribution of a *continuous* variable
  + but includes information like *median* (dark line) and *quartiles* (box and whiskers)
  + and *outliers* (dots)
- like scatterplots, require x and y variables
  + but one of them needs to be **categorical**

```{r, fig.cap="A scatterplot. Median (50th percentile): thick black lines; interquartile range (IQR; 25th and 75th percentile): box limits; minimum (0th percentile) and maximum (100th percentile) excluding outliers: : whiskers; outliers: points"}
#| output-location: column-fragment
iris |> 
  ggplot(aes(x = Species, y = Sepal.Length)) +
  labs(title = "A scatterplot",
       x = "Categorical variable",
       y = "Continuous variable") +
  geom_boxplot()
```

## Boxplot explained

::: {.content-visible when-format="revealjs"}
```{r echo = F, fig.align = "center"}
#| fig-cap: "Image source: @winter_statistics_2019 (all rights reserved)"
#| fig-cap-location: bottom

# invert colours for dark mode in slides
library(magick)
y <- magick::image_read(here::here("media/Winter_2019_boxplot.png"))

magick::image_negate(y)
```
:::

::: {.content-visible unless-format="revealjs"}
```{r echo = F, fig.align = "center"}
#| fig-cap: "Image source: @winter_statistics_2019 (all rights reserved)"
#| fig-cap-location: bottom
magick::image_read(here::here("media/Winter_2019_boxplot.png"))
```
:::

::: {.content-visible when-format="revealjs"}
::: notes
whiskers: 1.5*IQR from Q1 (lower whisker) or Q3 (upper whisker)
outliers: outside that range
:::
:::

## Boxplots

- let's change our scatterplot to a boxplot

```{r}
#| code-line-numbers: "4,"
#| output-location: column-fragment
df_lifetime |>
  filter(ff > 0,
         region == "verb") |>
  ggplot(aes(x = tense, y = ff)) +
  labs(title = "First-fixation duration (verb region)",
       x = "Tense",
       y = "First-fixation duration (ms)") +
  geom_boxplot(alpha = .2) +
  theme_bw()
```

### Grouped boxplots

```{r}
#| code-line-numbers: "4,"
#| output-location: column-fragment
df_lifetime |>
  filter(ff > 0,
         region == "verb") |>
  ggplot(aes(x = tense, y = ff, colour = lifetime)) +
  labs(title = "First-fixation duration (verb region)",
       x = "Tense",
       y = "First-fixation duration (ms)") +
  geom_boxplot(alpha = .2) +
  theme_bw()
```

#### Exercise

1. Create a group boxplot (x = tense, fill = lifetime) for
  + first-pass reading time (verb region)
  + regression path duration (verb region)
  + total reading time (verb region)
  + reaction times (use the `distinct()` verb to have a single observation per participant and per trial)

## Violin plots

## Violin boxplots


# Summary statistics

## Interaction plots {.smaller}

:::: {.columns}
::: {.column width="50%"}
- common for **factorial** designs, i.e., comparing *categorical* predictors
- there are 2 ways of producing them:
  + with your data frame and `stat_summary()`
  + or with a summary table and ggplot geoms `geom_point()`, `geom_errorbar()`, and `geom_line()`
- we'll need our summary table to plot an interaction plot
:::

::: {.content-hidden when-format="pdf"}
::: {.column width="50%"}
```{r}
#| echo: false
knitr::kable(summary_ff, digits = 1) |> 
  kableExtra::kable_styling(font_size = 20)
```
:::
:::

::: {.content-visible when-format="pdf"}
::: {.column width="50%"}
```{r}
#| echo: false

# smaller font for PDF output
knitr::kable(summary_ff, digits = 1)
```
:::
:::

::::

::: {.column width="50%"}
```{r}
#| echo: false
df_lifetime |> 
  filter(region == "verb") |> 
  ggplot(aes(x = lifetime, y = ff, 
                     shape = tense,
                     group = tense,
                     color = tense)) +
  labs(title="Interaction plot with `stat_summary()`",
       x = "Lifetime",
       y = "First fix (ms)",
       shape = "Tense", group = "Tense", color = "Tense", linetype = "Tense") +
  stat_summary(fun = "mean", geom = "point", size = 3, position = position_dodge(0.2)) +
  stat_summary(fun = "mean", geom = "line", position = position_dodge(0.2), aes(linetype=tense)) +
  stat_summary(fun.data = "mean_cl_normal", geom = "errorbar", width = .2
               , position = position_dodge(0.2)) +
  theme_bw()
```
:::

::: {.column width="50%"}
```{r}
#| echo: false
# plot the means and standard errors
summary_ff |> 
  ggplot(aes(x = lifetime, y = mean.ff, 
                     shape = tense,
                     group = tense,
                     color = tense)) +
  labs(title="Interaction plot with summary table and geoms",
       x = "Lifetime",
       y = "First fix (ms)",
       shape = "Tense", group = "Tense", color = "Tense", linetype = "Tense") +
  geom_point(size = 3,
                position = position_dodge(0.2)) +
  geom_line(aes(linetype=tense), position = position_dodge(0.2)) +
  geom_errorbar(aes(ymin = mean.ff - ci,
                    ymax = mean.ff + ci),
                width = .2,
                position = position_dodge(0.2)) +
  theme_bw()
```
:::

# Binomial data

- binomial data are those with 2 categories, for example
  + present, absent
  + yes, no
- in our dataset, each trial ended with a binary naturalness judgement task
  + how might we plot such data?
  
## Bar plot

::: {.column width="100%"}
- be sure to read in `accept` as a factor!
:::

::: {.column width="50%"}

```{r}
df_lifetime |> 
  distinct(px,trial,.keep_all=T) |> 
  ggplot(aes(x = as.factor(accept) )) +
  geom_bar() +
  theme_bw()
```

:::

::: {.column width="50%"}

```{r}
df_lifetime |> 
  distinct(px,trial,.keep_all=T) |> 
  ggplot(aes(x = as.factor(accept), fill = condition)) +
  labs(title = "Binary responses",
       x = "Naturalness response",
       fill = "Condition") +
  geom_bar() +
  theme_bw()
```

:::

## Grouped bar plots

```{r}
df_lifetime |> 
  distinct(px,trial,.keep_all=T) |> 
  ggplot(aes(x = as.factor(accept), fill = condition)) +
  labs(title = "Binary responses",
       x = "Naturalness response",
       fill = "Condition") +
  geom_bar(position = "dodge") +
  theme_bw()
```

### Exercise

1. Generate a grouped bar plot (i.e., `dodge`) with:
    - a facet grid for `tense`
    - plots `lifetime` on the x-axis
    - and fills the bars based on `accept`
    - change the labels accordingly
    - customise as you like
  
```{r, eval = F}
df_lifetime |> 
  distinct(px,trial,.keep_all=T) |> 
  ggplot(aes(x = lifetime, fill = as.factor(accept))) +
  facet_grid(.~tense) +
  labs(title = "Binary responses",
       x = "Lifetime",
       fill = "Response") +
  geom_bar(position = "dodge") +
  theme_bw()
```

## Grouped bar plots

```{r, eval = T}
df_lifetime |> 
  distinct(px,trial,.keep_all=T) |> 
  ggplot(aes(x = lifetime, fill = as.factor(accept))) +
  facet_grid(.~tense) +
  labs(title = "Grouped and faceted barplot",
       x = "Lifetime",
       fill = "Response") +
  geom_bar(position = "dodge") +
  theme_bw()
```

## Stacked bar plots

```{r, eval = T}
#| code-line-numbers: "8"
df_lifetime |> 
  distinct(px,trial,.keep_all=T) |> 
  ggplot(aes(x = lifetime, fill = as.factor(accept))) +
  facet_grid(.~tense) +
  labs(title = "Stacked and faceted barplot",
       x = "Lifetime",
       fill = "Response") +
  geom_bar(position = "stack") +
  theme_bw()
```

### Exercise

1. Choose the barplot you like best for binary data
2. Reproduce that barplot, but with `reg_in` at the `verb1` region

```{r}
#| echo: false
#| eval: true
#| code-line-numbers: "8"
df_lifetime |> 
  filter(region=="verb") |> 
  ggplot(aes(x = lifetime, fill = as.factor(reg_in))) +
  facet_grid(.~tense) +
  labs(title = "Regressions in (yes/no, verb region)",
       x = "Lifetime",
       fill = "Regressions in (yes/no)") +
  geom_bar(position = "stack") +
  theme_bw()
```

### Extra exercise

1. Create another bar plot, but for `reg_out` for all sentence regions
2. Use `facet_grid()`
  + to have facets by region (columns) and by tense (in 2 rows)

```{r}
#| echo: false
#| eval: true
#| code-line-numbers: "8"
df_lifetime |> 
  # filter(region=="verb") |> 
  ggplot(aes(x = lifetime, fill = as.factor(reg_out))) +
  facet_grid(tense~region) +
  labs(title = "Regressions in (yes/no, verb region)",
       x = "Lifetime",
       fill = "Regressions out (yes/no)") +
  geom_bar(position = "stack") +
  theme_bw()
```

# Resources

@nordmann_data_2022

@nordmann_applied_2022

@wickham_r_nodate, [Chapter 2](https://r4ds.hadley.nz/data-visualize.html)

# References {.unlisted .unnumbered visibility="uncounted"}

::: {#refs custom-style="Bibliography"}
:::




