---
title: "Multiple Regression"
subtitle: "Models with more than one predictor"
author: "Daniela Palleschi"
institute: Humboldt-Universität zu Berlin
footer: "Multiple Regression"
lang: en
date: "`r Sys.Date()`"
format:
  revealjs: 
    output-file: slides-mult_reg.html
    theme: [dark]
    width: 1600
    height: 900
    progress: true
    # smaller: true
    scrollable: true
    slide-number: c/t
    code-link: true
    code-overflow: wrap
    code-tools: true
    code-annotations: below
    # logo: logos/hu_logo.png
    # css: logo.css
    incremental: true
    # number-sections: true
    toc: true
    toc-depth: 1
    toc-title: 'Topics'
    navigation-mode: linear
    controls-layout: bottom-right
    fig-cap-location: top
    font-size: 0.6em
    slide-level: 4
    chalkboard: true
    title-slide-attributes: 
      data-background-image: logos/logos.tif
      data-background-size: 15%
      data-background-position: 50% 92%
  html:
    self-contained: true
    output-file: sheet-mult_reg.html
    theme: [dark]
    number-sections: true
    toc: true
    code-overflow: wrap
    code-tools: true
  pdf:
    output-file: pdf-mult_reg.pdf
    toc: true
    number-sections: false
    colorlinks: true
    code-overflow: wrap
editor_options: 
  chunk_output_type: console
bibliography: references/references.json
csl: references/apa.csl
execute:
  warning: false
  message: fal
---

```{r}
#| echo: false
## play sound if error encountered
### from: https://sejohnston.com/2015/02/24/make-r-beep-when-r-markdown-finishes-or-when-it-fails/
options(error = function(){    # Beep on error
  beepr::beep(sound = "wilhelm")
  Sys.sleep(2) # 
  }
 )
## and when knitting is complete
.Last <- function() {          # Beep on exiting session
  beepr::beep(sound = "ping")
  Sys.sleep(6) # allow to play for 6 seconds
  }
```

```{r, eval = T, cache = F}
#| echo: false
# Create references.json file based on the citations in this script
# make sure you have 'bibliography: references.json' in the YAML
# rbbt::bbt_update_bib("_mult_reg.qmd")
```

```{r}
#| echo: false
knitr::opts_chunk$set(eval = T, # change this to 'eval = T' to reproduce the analyses; make sure to comment out
                      echo = T, # 'print code chunk?'
                      message = F, # 'print messages (e.g., warnings)?'
                      error = F,
                      warning = F)
```

# Learning Objectives {.unnumbered}

Today we will learn...

- what multiple regression is
- fit a model with multiple predictors
- learn to interpret interactions

# Set-up environment  {.unnumbered}

```{r}
# suppress scientific notation
options(scipen=999)
```

## Packages {.unnumbered}

Run this in the Console:

```{r}
#| eval: false
install.packages("devtools")
devtools::install_github("strengejacke/sjPlot")
```

Then copy this into your script:

```{r}
# load libraries
pacman::p_load(
  tidyverse,
  here,
  knitr,
  kableExtra,
  gt,
  broom,
  Rmisc,
  patchwork,
  sjPlot
)
```




## Data {.unnumbered}

```{r}
# read in data
df_crit_verb <- readr::read_csv(here::here("data/tidy_data_lifetime_pilot.csv"), 
                               # for special characters
                               locale = readr::locale(encoding = "latin1") 
                               ) |> 
  mutate_if(is.character,as.factor) |> # all character variables as factor
  filter(type == "critical", # only critical trials
         px != "px3",
         region == "verb") |>
  droplevels()
```

# Resources {.unnumbered}

Chapters 6 and 7 in @winter_statistics_2019

# Review: Linear Regression

- let's re-run our simple linear model: `fp ~ lifetime`

## Set contrasts

Let's stick to sum contrast coding (+/-0.5).

```{r}
# order factor levels
df_crit_verb$lifetime <- factor(df_crit_verb$lifetime, levels = c("living","dead"))
# or using the tidyverse
df_crit_verb <-
  df_crit_verb %>% 
  mutate(lifetime = fct_relevel(lifetime, "living", "dead"))
```

```{r}
# set contrasts
contrasts(df_crit_verb$lifetime) <- c(-0.5,+0.5)
```

```{r}
contrasts(df_crit_verb$lifetime)
```

## Fit model

```{r}
# fit simple linear model
fit_fp_lifetime <- df_crit_verb %>%
  filter(fp > 0) %>%
  lm(fp ~ lifetime, data = .)
```

```{r}
# check our contrasts
coef(fit_fp_lifetime)
```

## Lifetime Effect

- recall that we found a significant effect of `lifetime`
  + the `dead` conditions elicited longer first-pass reading times than the `living` conditions
- but we also had two tenses: the Present Perfect (`PP`) and the Simple Future (`SF`)
  + how can we also check for an effect of `tense`?
  
## Another linear model

- we could run a second model with `tense` as the predictor instead of `lifetime`

```{r}
df_crit_verb <-
  df_crit_verb %>% 
  mutate(tense = fct_relevel(tense, "PP", "SF"))
```

```{r}
contrasts(df_crit_verb$tense) <- c(-0.5,+0.5)
```

```{r}
contrasts(df_crit_verb$tense)
```

```{r}
# fit simple linear model
fit_fp_tense <-
  df_crit_verb %>%
  filter(fp > 0) %>%
  lm(fp ~ tense, data = .)
```

---

- what does the summary tell us?

```{r}
summary(fit_fp_tense)
```

# Multiple regression

- a simple linear model is essentially just modelling the mean
- so is multiple regression, but takes more conditions into account:

$$
y = b_0 + b_1x + b_2x + ... + e
$$

- the slope index ($b_1$, $b_2$) differentiates between predictors

$$
fp = b_0 + (b_1\times lifetime) + (b_2\times tense) + ... + e
$$

## Adding another predictor

- in our experimental design we were interested in the effect of *lifetime* on the processing of two different *tenses*: the Present Perfect and Simple Future
  + in other words, we had two fixed factors: lifetime and tense

- multiple regression is not the same as running separate simple linear models, rather each coefficient becomes a *partial* regression coefficient

## Fit model

+ we can use `+` to indicate we have another fixed effect, and that we want to look at the **main effect** of this predictor

```{r}
# multiple regression
fit_fp <-
  df_crit_verb %>%
  filter(fp > 0) %>% 
  lm(fp ~ lifetime + tense, data = .)
```

### Model summary

```{r}
summary(fit_fp)
```

## Comparing summaries {.smaller}


```{r}
#| label: tbl-lifetime
#| tbl-cap: Coefficients for fit_fp_lifetime (fp ~ lifetime)
tidy(fit_fp_lifetime) %>% kable() %>% kable_styling()
```

```{r}
#| label: tbl-tense
#| tbl-cap: Coefficients for fit_fp_tense (fp ~ tense)
tidy(fit_fp_tense) %>% kable() %>% kable_styling()
```

```{r}
#| label: tbl-mlm
#| tbl-cap: Coefficients for fit_fp (fp ~ lifetime + tense)
tidy(fit_fp) %>% kable() %>% kable_styling()
```


## Partial regression coefficients

- in multiple regression, each predictor becomes a *partial* regression coefficient
  + the effect of one predictor while holding all other predictors constant
  
$$
\begin{align}
fp &= 309.1 + (31.5\times lifetime) + (-12.8\times tense)\\
fp_{dead-PP} &= 309.1 + (31.5\times 0.5) + (-12.8\times -0.5)
\end{align}
$$

## Exploring the model: fixed effects {-}

- what were our coefficients?
```{r}
#| output-location: fragment
coef(fit_fp)
```

- what is the mean of `lifetime` coded as `-0.5` (`living`)?
```{r}
#| output-location: fragment
coef(fit_fp)['(Intercept)'] + coef(fit_fp)['lifetime1'] * -0.5
```

- ignore the `(Intercept)` label here, `R` just takes the first label when performing an operation on 2 vectors

- what is the mean of `lifetime` coded as `+0.5` (`dead`)?
```{r}
#| output-location: fragment
coef(fit_fp)['(Intercept)'] + coef(fit_fp)['lifetime1'] * 0.5
```

### Exercise

- do the same for `tense`:
  - find the fitted value for `PP` and for `SF`

## Exploring the model: predicting fp

:::: columns

::: {.column width="50%"}
```{r}
#| echo: false
#| label: tbl-estimates
#| tbl-cap: Coefficient estimates
tidy(fit_fp) %>% 
  select(term,estimate) %>% 
  mutate(`*0.5` = estimate*.5) %>% 
  kable(digits = 1) %>% 
  kable_styling()
  # fmt_number(decimals = 5)
```
:::
::: {.column width="50%"}
```{r}
#| echo: false
#| label: tbl-fitted
#| tbl-cap: Fitted values per condition
augment(fit_fp, data = df_crit_verb[df_crit_verb$fp > 0,]) %>% 
  dplyr::summarise(fp = mean(fp),
            .by = c(lifetime,tense,.fitted)) %>% 
  mutate(mean_diff = fp-.fitted) %>% 
  arrange(lifetime) %>% 
  kable(digits = 1) %>% 
  kable_styling()
```
:::
::::

### Calculating fitted values by hand

$$
fp = b_0 + (b_1\times lifetime) + (b_2\times tense) + e
$$

```{r}
# living-PP
coef(fit_fp)['(Intercept)'] + 
  coef(fit_fp)['lifetime1'] * -0.5 + coef(fit_fp)['tense1'] * -0.5 
```
```{r}
# living-SF
coef(fit_fp)['(Intercept)'] + 
  coef(fit_fp)['lifetime1'] * -0.5 + coef(fit_fp)['tense1'] * 0.5 
```
```{r}
# dead-PP
coef(fit_fp)['(Intercept)'] + 
  coef(fit_fp)['lifetime1'] * 0.5 + coef(fit_fp)['tense1'] * -0.5 
```
```{r}
# dead-SF
coef(fit_fp)['(Intercept)'] + 
  coef(fit_fp)['lifetime1'] * 0.5 + coef(fit_fp)['tense1'] * 0.5 
```


## Exploring the model: residuals {-}

- recall that residuals = the difference between our *fitted* (model) values and our *observed* (data) values

```{r}
#| output-location: fragment
# what do our FITTED values look like?
head(fitted(fit_fp))
```

```{r}
#| output-location: fragment
# what do our OBSERVED values look like?
head(df_crit_verb$fp)
```

```{r}
#| output-location: fragment
# what is the difference between the FITTED and OBSERVED values?
head(df_crit_verb$fp) - head(fitted(fit_fp))
```

```{r}
#| output-location: fragment
# what are our RESIDUALS?
head(residuals(fit_fp))
```



## Plotting our effects

- based on these plots, it seems like the effects of `lifetime` and `tense` differed
  + the slant in figure A (effect of `lifetime`) looks steeper for the `PP` than the `SF`
- this could reflect an interaction effect

```{r}
#| echo: false
fig_error_lifetime <-
  df_crit_verb %>% 
  filter(fp > 0) %>% 
  summarySEwithin(measurevar="fp", withinvars=c("lifetime"), idvar="px") %>% 
  mutate(upper = fp+ci,
         lower = fp-ci) %>% 
  ggplot(aes(x = lifetime, y = fp, colour = lifetime)) + 
  labs(title = "Mean first-pass reading times (with 95% CIs)") +
  geom_point(position = position_dodge(0.2), size = 2) +
  geom_line(group=1,position = position_dodge(0.2), colour = "grey") +
  geom_errorbar(aes(ymin=lower,ymax=upper), position = position_dodge(0.2), width = .2) +
  theme_bw() +
  theme(text = element_text(size=8))

fig_scatter_lifetime <-
  df_crit_verb %>% 
  filter(fp > 0) %>% 
  ggplot(aes(x = lifetime, y = fp, colour = lifetime, shape = lifetime, fill = lifetime)) + 
  labs(title = "Distribution of first-pass reading times") +
  geom_point(position = position_jitterdodge(jitter.width = 0.3), size = 2, alpha = .5) +
  geom_violin(alpha = .4) +
  theme_bw()+
  theme(text = element_text(size=8))

```

```{r}
#| echo: false
fig_error_tense <-
  df_crit_verb %>% 
  filter(fp > 0) %>% 
  summarySEwithin(measurevar="fp", withinvars=c("tense"), idvar="px") %>% 
  mutate(upper = fp+ci,
         lower = fp-ci) %>% 
  ggplot(aes(x = tense, y = fp, colour = tense)) + 
  labs(title = "Mean first-pass reading times (with 95% CIs)") +
  geom_point(position = position_dodge(0.2), size = 2) +
  geom_line(group=1,position = position_dodge(0.2), colour = "grey") +
  geom_errorbar(aes(ymin=lower,ymax=upper), position = position_dodge(0.2), width = .2) +
  theme_bw() +
  theme(text = element_text(size=8))

fig_scatter_tense <-
  df_crit_verb %>% 
  filter(fp > 0) %>% 
  ggplot(aes(x = tense, y = fp, colour = tense, shape = tense, fill = tense)) + 
  labs(title = "Distribution of first-pass reading times") +
  geom_point(position = position_jitterdodge(jitter.width = 0.3), size = 2, alpha = .5) +
  geom_violin(alpha = .4) +
  theme_bw()+
  theme(text = element_text(size=8))
```


```{r}
#| echo: false
#| label: fig-main
#| fig-cap: Plotting our two predictors
#| out-width: "100%"
(fig_error_lifetime + fig_scatter_lifetime) /
  (fig_error_tense + fig_scatter_tense) + plot_annotation(tag_levels = "A")
```

```{r}
#| eval: false
#| echo: false
augment(fit_fp, data = df_crit_verb[df_crit_verb$fp > 0,]) %>% 
  dplyr::summarise(fp = mean(fp),
            .by = c(lifetime,tense,.fitted)) %>% 
  arrange(lifetime)  %>% 
  ggplot(aes(x = lifetime, y = .fitted, colour = tense)) +
  geom_point(position = position_dodge(0.2)) +
  geom_line(aes(group=tense), position = position_dodge(0.2))
  
```

### Plotting coefficients with `sjPlot`

```{r}
plot_model(fit_fp) +
  geom_hline(yintercept=0) +
  theme_bw()
```



# Interactions

- interactions are when the effect of a predictor on an outcome depends on another predictor
  + an example from @mcelreath_statistical_nodate: plant growth cannot be predicted solely by sun exposure or water exposure, but rather the two
  + the influence of the sun is dependent on water, and vice versa

- the equation for multiple regression with an interaction effect is as follows:

$$
y = b_0 + b_1x_1 + b_2x_2 + b_3(x_1*x_2) + e
$$

- where $b_0, b_1, b_2$ equal the intercept, and the slopes ($b_1, b_2$) of two variables ($x_1, x_2$)
- the interaction ($b_3$) is the multiplication of $x_1$ and $x_2$

## Fitting an interaction term to our model

- to add an interaction, we can use a colon to separate two variables
  + `lifetime:tense`
- however, it's common to use an asterisk instead, which means "fit main effects of these variables *and* their interaction*
  + `lifetime:tense`
  
```{r}
# multiple regression
fit_fp_inter <-
  df_crit_verb %>%
  filter(fp > 0) %>% 
  lm(fp ~ lifetime*tense, data = .)
```

```{r}
#| output-location: slide
summary(fit_fp_inter)
```

## Plotting an interaction


```{r}
#| echo: false
fig_error <-
  df_crit_verb %>% 
  filter(fp > 0) %>% 
  summarySEwithin(measurevar="fp", withinvars=c("lifetime", "tense"), idvar="px") %>% 
  mutate(upper = fp+ci,
         lower = fp-ci) %>% 
  ggplot(aes(x = lifetime, y = fp, colour = tense, shape = tense)) + 
  labs(title = "Mean first-pass reading times (with 95% CIs)") +
  geom_point(position = position_dodge(0.2), size = 2) +
  geom_line(position = position_dodge(0.2), aes(group=tense)) +
  geom_errorbar(aes(ymin=lower,ymax=upper), position = position_dodge(0.2), width = .2) +
  theme_bw() +
  theme(text = element_text(size=8))

fig_scatter <-
  df_crit_verb %>% 
  filter(fp > 0) %>% 
  ggplot(aes(x = tense, y = fp, colour = lifetime, shape = lifetime, fill = lifetime)) + 
  labs(title = "Distribution of first-pass reading times") +
  geom_point(position = position_jitterdodge(jitter.width = 0.3), size = 2, alpha = .5) +
  introdataviz::geom_split_violin(alpha = .4) +
  theme_bw()+
  theme(text = element_text(size=8))

```



```{r}
#| echo: false
#| label: fig-inter
#| fig-cap: Plotting our two predictors
#| out-width: "100%"
fig_error + fig_scatter + plot_annotation(tag_levels = "A")
```

### Plotting an interaction with `sjPlot`

```{r}
fig_sjplot_coef <- plot_model(fit_fp_inter) +
  geom_hline(yintercept=0) +
  theme_bw()

fig_sjplot_int <- plot_model(fit_fp_inter, type = "int") +
  geom_hline(yintercept=0) +
  geom_line(group=1) +
  theme_bw()
```

```{r}
#| label: fig-sjplot-int
#| fig-cap: Plotting model coefficients (A) and  interactions (B) with the sjPlot package
fig_sjplot_coef + fig_sjplot_int + plot_annotation(tag_levels = "A")
```


# Exercise

1.Run a model with main and interaction effects for `tense` and `lifetime` on `total reading time` at the `verb` region.

2. Print the summary and try to interpret it.

3. Plot the main and interaction effects, and the coefficients.




# References {.unlisted .unnumbered visibility="uncounted"}

::: {#refs custom-style="Bibliography"}
:::
